```{r}
#| label: element-context
#| include: false
# Everything below is driven by SYM, set by the calling page. No page holds a
# number of its own: they all read the same summary CSVs.
e_rows <- elements[elements$symbol == SYM, ]
e_rows <- e_rows[order(match(e_rows$cat, cat_order)), ]
e_name <- e_rows$name[1]
e_cap  <- capability(SYM)
e_bulk <- e_rows[e_rows$cat == "bulk", ]
has_bulk <- nrow(e_bulk) > 0

e_bg  <- backgrounds[backgrounds$symbol == SYM, ]
e_vd  <- verdicts[verdicts$symbol == SYM, ]
e_pr  <- pressure[pressure$symbol == SYM, ]
e_ct  <- controls[controls$symbol == SYM & controls$reliable, ]
# A withheld element gets no distance-band chart. Its surviving concentrations are
# an upper tail, so a median by band would be a number we already know is biased,
# drawn as if it were a reading.
show_pressure <- !any(e_rows$withheld) &&
  nrow(e_pr[e_pr$measure == "concentration" & e_pr$reliable, ]) > 0
show_controls <- !any(e_rows$withheld) && nrow(e_ct) > 0
```

```{r}
#| output: asis
keyfig(
  list(format(e_rows$n_elem[1], big.mark = "\u00a0"), "measurements"),
  list(format(e_rows$n_sites_elem[1], big.mark = "\u00a0"), "sites"),
  list(paste0(e_rows$year_min_elem[1], " to ", e_rows$year_max_elem[1]), "years"),
  list(paste0(e_rows$n_sources_elem[1], " of ", nrow(sources)), "sources")
)
cat(chip(e_cap), " ", capability_note(SYM), "\n\n", sep = "")
```

## How much data there is

```{r}
e_rows |>
  transmute(Fraction = cat_label[cat],
            Measurements = format(n, big.mark = "\u00a0"),
            Sites = format(n_sites, big.mark = "\u00a0"),
            Datasets = n_datasets,
            Sources = n_sources,
            Years = ifelse(is.na(year_min), "-", paste0(year_min, " to ", year_max))) |>
  kable()
```

```{r}
#| output: asis
#| eval: !expr any(e_rows$withheld)
cat("## Why this element is withheld\n\n",
    "The laboratories that measured ", tolower(e_name), " often could not quantify it: ",
    "the value was somewhere below a detection limit rather than a number. Those rows ",
    "are removed during cleaning, which is right for contamination screening and wrong ",
    "for estimating a background.\n\n",
    "A background **is** the low end of a distribution. Delete the low end, and what ",
    "is left is an upper tail that would look like a respectable background and be too ",
    "high. So no background and no verdict is published for ", tolower(e_name),
    ", at any fraction.\n\n", sep = "")
```

```{r}
#| eval: !expr any(e_rows$withheld)
#| fig-height: 2.6
#| fig-cap: !expr sprintf("Share of %s measurements that fell below the limit of quantification, by source. The dashed line is the 20 per cent limit above which a background is withheld.", tolower(e_name))
censoring |>
  filter(symbol == SYM) |>
  mutate(source = ifelse(source == "ALL", "All sources", source),
         source = reorder(source, pct_censored)) |>
  ggplot(aes(pct_censored, source, fill = source == "All sources")) +
  geom_col() +
  geom_vline(xintercept = 20, linetype = 2, colour = "#b3261e") +
  geom_text(aes(label = sprintf("%.1f%%", pct_censored)), hjust = -0.15, size = 3.1) +
  scale_fill_manual(values = c(`FALSE` = "#bbbbbb", `TRUE` = "#CC6677"), guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.22))) +
  labs(x = "Below the limit of quantification (%)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank())
```

```{r}
#| output: asis
#| eval: !expr any(e_rows$withheld)
cat("The concentrations themselves are still in the database and the downloadable ",
    "dataset, flagged as withheld. What is withheld is the *interpretation*, not the ",
    "data.\n\n", sep = "")
```

```{r}
#| output: asis
#| eval: !expr isFALSE(any(e_rows$has_background)) && isFALSE(any(e_rows$withheld))
cat("## Why there is no result\n\n",
    "There is simply not enough of it. With fewer measurements than the reporting ",
    "threshold of ", smeta$min_n, ", any background would be an artefact of the handful ",
    "of samples that happen to exist, and would carry a false air of authority. The ",
    "measurements are published in the dataset; no estimate is built on them.\n\n",
    sep = "")
```

```{r}
#| output: asis
#| eval: !expr any(e_rows$has_background)
cat("## Background levels\n\n",
    "Five ways of estimating the natural level, in ", UNIT,
    ". A blank means the method could not be applied.\n\n", sep = "")
```

```{r}
#| eval: !expr any(e_rows$has_background)
# igeo_b is deliberately not shown: the Igeo reference IS the offshore median, so
# it would be a second identical column. It stays in summary_background.csv.
lvl <- c("offshore", "offshore_p90", "mixture", "gsnorm", "pressure_far")
lab <- c(offshore = "Offshore median", offshore_p90 = "Offshore P90",
         mixture = "Mixture bound", gsnorm = "Grain-size normalised",
         pressure_far = "Far from farms")
e_bg |>
  filter(method %in% lvl) |>
  mutate(shown = ifelse(reliable, fmt(value), NA_character_),
         Fraction = factor(cat_label[cat], levels = cat_label[cat_order]),
         method = factor(lab[method], levels = lab[lvl])) |>
  select(Fraction, method, shown) |>
  arrange(Fraction, method) |>
  pivot_wider(names_from = method, values_from = shown, names_sort = FALSE) |>
  kable()
```

```{r}
#| output: asis
#| eval: !expr any(e_rows$has_verdict)
cat("## Pristine verdict\n\n",
    "A sample is judged against its own grain size, using aluminium. ",
    "The permissive rule is a single test; the strict rule requires three ",
    "independent criteria to agree.\n\n",
    "Only bulk sediment is listed. A sieved sample has already been grain-size ",
    "corrected by the sieve, so aluminium has nothing left to correct: see ",
    "[Estimation of Pristine Sediment](methods-pristine.qmd#and-in-bulk-sediment-only)",
    ".\n\n", sep = "")
```

```{r}
#| eval: !expr any(e_rows$has_verdict)
e_vd |>
  filter(reliable) |>
  transmute(Fraction = cat_label[cat],
            Measurements = format(n, big.mark = "\u00a0"),
            `Can be judged` = sprintf("%s%% (%s)", pct_classifiable,
                                      format(n_classifiable, big.mark = "\u00a0")),
            `Pristine, permissive` = sprintf("%s%%", pct_ef),
            `Pristine, strict` = sprintf("%s%%", pct_strict)) |>
  kable()
```

```{r}
#| output: asis
#| eval: !expr any(!is.na(e_vd$pct_igeo_class0))
cat("## Contamination classes\n\n",
    "The geo-accumulation index compares each sample with a regional background ",
    "concentration. It needs no aluminium, so it covers far more of the data than ",
    "the pristine verdict does, at the cost of not controlling for grain size. Its ",
    "reference is the offshore median in the table above.\n\n",
    sep = "")
```

```{r}
#| eval: !expr any(!is.na(e_vd$pct_igeo_class0))
e_vd |>
  filter(!is.na(pct_igeo_class0)) |>
  transmute(Fraction = cat_label[cat],
            Measurements = format(n_igeo, big.mark = "\u00a0"),
            `Unpolluted (Igeo at or below 0)` = sprintf("%s%%", pct_igeo_class0)) |>
  kable()
```

```{r}
#| output: asis
#| eval: !expr show_pressure
cat("## Distance to the nearest fish farm\n\n")
```

```{r}
#| eval: !expr show_pressure
#| fig-height: 3
#| fig-cap: !expr sprintf("Median %s concentration by distance to the nearest marine fish farm, %s.", tolower(e_name), UNIT)
e_pr |>
  filter(measure == "concentration", reliable) |>
  mutate(band = factor(band, levels = c("<1km", "1-5km", "5-20km", ">20km")),
         Fraction = cat_label[cat]) |>
  ggplot(aes(band, value, fill = Fraction)) +
  geom_col(position = position_dodge(width = .8), width = .7) +
  scale_fill_manual(values = c(Bulk = "#4477AA", `Sieved <63 um` = "#66CCEE",
                               `Sieved <20 um` = "#228833")) +
  labs(x = "Distance to the nearest fish farm", y = paste0("Median (", UNIT, ")"),
       fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank(), legend.position = "top")
```

```{r}
#| output: asis
#| eval: !expr show_controls
cat("Raw distance bands are confounded: the far band contains monitored pollution, ",
    "and it is muddier than sediment under a cage. The controlled comparisons are ",
    "the ones to read. See [Aquaculture Pressure](methods-pressure.qmd).\n\n", sep = "")
```

```{r}
#| eval: !expr show_controls
e_ct |>
  transmute(Control = control,
            `What it holds constant` = holds,
            `Samples near / far` = sprintf("%s / %s", format(n_near, big.mark = "\u00a0"),
                                           format(n_far, big.mark = "\u00a0")),
            `Near vs far` = detail) |>
  kable()
```

## Where the samples are

```{r}
#| include: false
grid_path <- file.path(sdir, "summary_map_grid.csv")
g <- read.csv(grid_path, stringsAsFactors = FALSE)
g <- g[g$symbol == SYM & g$cat == "bulk",
       c("lat", "lon", "n", "value_p50", "igeo_p50", "pct_below_p90")]
if (!nrow(g)) g <- g[0, ]
ojs_define(mapdata = g)
map_has_points <- nrow(g) > 0
# Igeo needs a background, and a withheld element has none, so the grid carries no
# Igeo for it. Offering a colour scale that would paint every cell grey is worse
# than not offering it.
map_has_igeo <- map_has_points && any(!is.na(g$igeo_p50))
ojs_define(map_unit = UNIT)
ojs_define(map_has_igeo = map_has_igeo)
```

```{r}
#| output: asis
#| eval: !expr isFALSE(map_has_points)
cat("There is no bulk-sediment map for this element: too few measurements.\n")
```

```{=html}
<link rel="stylesheet" href="https://unpkg.com/leaflet@1.9.4/dist/leaflet.css" />
```

```{ojs}
//| echo: false
//| eval: true
L = require("https://unpkg.com/leaflet@1.9.4/dist/leaflet.js")
pts = transpose(mapdata)
```

```{ojs}
//| echo: false
viewof chosen = map_has_igeo
  ? Inputs.radio(
      new Map([["Concentration", "value_p50"], ["Enrichment (Igeo)", "igeo_p50"]]),
      { value: "value_p50", label: "Colour by" })
  : html`<span></span>`
layer = map_has_igeo ? chosen : "value_p50"
```

```{ojs}
//| echo: false
mapDiv = {
  const div = html`<div id="map"></div>`;
  yield div;
  const map = L.map(div, { zoomControl: true });
  // CARTO's basemaps now watermark every tile with "API key required", so the
  // tiles come straight from OpenStreetMap, as they do on the detailed site.
  L.tileLayer("https://tile.openstreetmap.org/{z}/{x}/{y}.png", {
    attribution: "&copy; OpenStreetMap contributors",
    maxZoom: 12
  }).addTo(map);

  // The home view is this element's own extent, not a fixed window on Norway.
  // Copper runs from 36 to 81 north; iodine sits in the Kattegat. The zoom cap
  // stops a sparse element from opening at street level with no coastline for
  // context, and the fallback is the extent of the database as a whole.
  const lats = pts.map(d => d.lat), lons = pts.map(d => d.lon);
  const home = pts.length
    ? L.latLngBounds([Math.min(...lats), Math.min(...lons)],
                     [Math.max(...lats), Math.max(...lons)])
    : L.latLngBounds([36, -25], [81, 40]);
  // Fit to the data, then hold the zoom inside 3 to 6. The cap stops a sparse
  // element opening at street level with no coastline for context. The floor is
  // there because Mercator stretches the far north: fitting copper's 36 to 81
  // degrees honestly would open on a world map with Europe as a smudge in the
  // middle. Three is the zoom the detailed site's maps use.
  const goHome = () => {
    map.invalidateSize();
    map.fitBounds(home, { padding: [24, 24], maxZoom: 6 });
    if (map.getZoom() < 3) map.setView(home.getCenter(), 3);
  };
  goHome();

  const HomeControl = L.Control.extend({
    options: { position: "topleft" },
    onAdd() {
      const btn = L.DomUtil.create("a", "leaflet-home-button");
      btn.href = "#";
      btn.title = "Reset the view";
      btn.setAttribute("aria-label", "Reset the view");
      btn.innerHTML = '<i class="bi bi-house-door" aria-hidden="true"></i>';
      L.DomEvent.on(btn, "click", e => { L.DomEvent.stop(e); goHome(); });
      return btn;
    }
  });
  new HomeControl().addTo(map);

  const vals = pts.map(d => d[layer]).filter(v => v !== null && !isNaN(v));
  if (!vals.length) return;
  const sorted = vals.slice().sort((a, b) => a - b);
  const q = p => sorted[Math.floor(p * (sorted.length - 1))];
  const lo = q(0.05), hi = q(0.95);
  // Diverging around zero for Igeo, sequential for concentration. Igeo is a log2
  // index whose zero point is "at background", so a scale that does not mark zero
  // would hide the only threshold that means anything.
  const ramp = layer === "igeo_p50"
    ? d3.scaleDiverging(d3.interpolateRdBu).domain([Math.max(2, hi), 0, Math.min(-2, lo)])
    : d3.scaleSequential(d3.interpolateViridis).domain([lo, hi]);

  const canvas = L.canvas({ padding: 0.5 });
  for (const d of pts) {
    const v = d[layer];
    if (v === null || isNaN(v)) continue;
    L.circleMarker([d.lat, d.lon], {
      renderer: canvas, radius: 4, weight: 0.4, color: "#333",
      fillColor: ramp(v), fillOpacity: 0.85
    }).bindPopup(
      `<b>${d.lat.toFixed(1)}, ${d.lon.toFixed(1)}</b><br>` +
      `median: ${d.value_p50} ${map_unit}<br>` +
      `Igeo: ${d.igeo_p50 ?? "n/a"}<br>` +
      `measurements: ${d.n}`
    ).addTo(map);
  }
}
```

Each point is a 0.1 degree cell (roughly 11 km), coloured by the median of the
sites inside it. Cells are summaries: a cell containing both clean and enriched
sites shows the middle of them. The map opens on the region where this element was
measured, which differs from element to element; the house button returns to that
view.

Bulk sediment only. The sieved fractions are measured and published, but they are a
different material and cannot be drawn on the same colour scale.
