```{r}
#| label: source-context
#| include: false
# Everything below is driven by SRC, the source key set by the calling page. As
# everywhere on this site, no page holds a number of its own: these all come from
# summary_source_*.csv, written by analyze_data("refined", module = "summary").
s_lab <- unname(src_levels[SRC])
s_row <- sources[sources$source == s_lab, ]
s_el  <- source_elements[source_elements$key == SRC, ]
s_el  <- s_el[order(match(s_el$symbol, elem_levels)), ]
s_fl  <- source_flow[source_flow$key == SRC, ]
s_fl  <- s_fl[order(match(s_fl$stage, names(stage_label))), ]
s_rm  <- source_removals[source_removals$key == SRC, ]
s_rm  <- s_rm[order(match(s_rm$reason, removal_order)), ]
s_fr  <- source_fractions[source_fractions$key == SRC, ]
s_fr  <- s_fr[order(match(s_fr$cat, frac_order)), ]
s_ex  <- source_extras[source_extras$key == SRC, ]

s_has    <- s_el[s_el$in_archive, ]
s_missing <- s_el[!s_el$in_archive, ]
n_pilot <- s_fl$n[s_fl$stage == "pilot"]
n_slim  <- s_fl$n[s_fl$stage == "slim"]

# a list that reads as English, whether it holds one name or five
and_list <- function(x) if (length(x) < 2) paste(x) else
  paste(paste(utils::head(x, -1), collapse = ", "), "and", utils::tail(x, 1))
# Element names are stored in title case, which is right in a table and wrong in
# the middle of a sentence.
sentence <- function(x) sub("^(.)", "\\U\\1", x, perl = TRUE)
```

```{r}
#| output: asis
keyfig(
  list(format(s_row$n, big.mark = " "), "measurements reported"),
  list(format(s_row$n_sites, big.mark = " "), "sites"),
  list(sprintf("%d of %d", nrow(s_has), nrow(s_el)), "target elements carried"),
  list(paste0(s_row$year_min, " to ", s_row$year_max), "sampling years")
)
```

## What this archive is

```{r}
#| output: asis
cat(src_blurb[[SRC]], "\n\n")
cat(sprintf(paste("The parse of this archive, before anything was harmonised, has",
                  "a site of its own: [%s](%s)."),
            paste0(SRC, "-pilot"), src_pilot(SRC)), "\n")
```

## Which elements it carries

```{r}
s_el |>
  transmute(Element = sprintf("%s (%s)", name, sym_display[as.character(symbol)]),
            `In the archive` = ifelse(in_archive, format(n_slim, big.mark = " "), "-"),
            `Reported here` = ifelse(in_archive, format(n_analysed, big.mark = " "), "-"),
            Sites = ifelse(in_archive, format(n_sites, big.mark = " "), "-"),
            Years = ifelse(in_archive & !is.na(year_min),
                           paste0(year_min, " to ", year_max), "-"),
            `Below LOQ` = ifelse(is.na(pct_censored), "-",
                                 sprintf("%.1f%%", pct_censored))) |>
  kable(align = c("l", rep("r", 5)))
```

**In the archive** is what the harmonised database holds for this source before
any quality rule was applied. **Reported here** is what survives into the figures
on this site. **Below LOQ** is the share of the archive's measurements that fell
under the laboratory's limit of quantification and were deleted, which is the
single biggest reason the two columns differ.

```{r}
#| output: asis
if (nrow(s_missing)) {
  miss <- elem_name[as.character(s_missing$symbol)]
  cat(sprintf("This archive carries no measurement of %s at all. That is a property of the archive and not something the pipeline did: %s never reported into it in the first place.\n\n",
              tolower(and_list(miss)),
              if (length(miss) == 1) "it was" else "they were"))
}
wh <- s_has[s_has$withheld, ]
if (nrow(wh)) {
  cat(sprintf("%s %s carried here but withheld across the whole study, because too much of the data sits below the detection limit for any average to mean anything. See [All elements at a glance](results.qmd).\n",
              sentence(tolower(and_list(elem_name[as.character(wh$symbol)]))),
              if (nrow(wh) == 1) "is" else "are"))
}
```

## Which fraction of the sediment it measured

Grain size is the largest single control on a trace-element concentration: the
fine grains carry most of the metal, so a sieved sample and an unsieved one taken
from the same seabed are not the same measurement. Which fraction an archive works
in is part of what it is, and it is why this site reports bulk and the two
standard sieved fractions separately instead of pooling them.

```{r}
s_fr |>
  filter(cat != "other" | n > 0) |>
  transmute(Fraction = frac_label[cat],
            Measurements = ifelse(n > 0, format(n, big.mark = " "), "-"),
            Share = ifelse(n > 0, sprintf("%.1f%%", pct_of_source), "-"),
            Sites = ifelse(n > 0, format(n_sites, big.mark = " "), "-"),
            Elements = ifelse(n > 0, as.character(n_elements), "-"),
            Years = ifelse(n > 0 & !is.na(year_min),
                           paste0(year_min, " to ", year_max), "-"),
            `What it is` = what) |>
  kable(align = c("l", rep("r", 5), "l"))
```

```{r}
#| output: asis
# Which fraction this archive works in is a fact about the archive, so it is read
# off the shares rather than asserted per source in the text.
std <- s_fr[s_fr$cat != "other", ]
std <- std[order(-std$n), ]
top <- std[1, ]
rest <- std[std$n > 0, ][-1, , drop = FALSE]
cat(if (top$pct_of_source >= 99)
      sprintf("Effectively everything this archive reports is %s, so nothing on this page mixes fractions.\n\n",
              tolower(frac_label[top$cat]))
    else sprintf("Most of it is %s (%.0f in every hundred), and it also reports %s. They are kept apart everywhere on this site, because a background estimated on one fraction does not transfer to another.\n\n",
                 tolower(frac_label[top$cat]), top$pct_of_source,
                 tolower(and_list(unname(frac_label[rest$cat])))))

oth <- s_fr[s_fr$cat == "other", ]
if (nrow(oth) && oth$n > 0)
  cat(sprintf("A further %s measurements (%.1f%%) sit in a fraction this study does not pool with the three above. They are in the archive and in the database, and they are not reported on this site, because there are too few of them anywhere to estimate anything from.\n\n",
              format(oth$n, big.mark = " "), oth$pct_of_source))

cat("A site sampled in two fractions is counted in both rows, so the site column is not a total.\n")
```

## Aluminium, iron and organic carbon

None of these three is something this study measures. They are what makes a
measurement readable: aluminium is the grain-size normaliser an enrichment factor
is built on, iron carries much the same signal but can never enter a verdict here,
and organic carbon binds metals and is read beside a concentration. Coverage is
the whole question, so coverage is all that is shown.

```{r}
s_ex |>
  transmute(Parameter = name,
            `In the archive` = ifelse(in_archive, format(n_slim, big.mark = " "), "-"),
            Sites = ifelse(in_archive, format(n_sites, big.mark = " "), "-"),
            `Measurements paired` = ifelse(in_archive,
                                           format(n_paired, big.mark = " "), "-"),
            Share = ifelse(in_archive, sprintf("%.1f%%", pct_paired), "-"),
            `What it is for` = what) |>
  kable(align = c("l", rep("r", 4), "l"))
```

**In the archive** counts the archive's own measurements of that parameter.
**Measurements paired** counts something else: the target measurements on this site
that have one of them beside them, on the same sample and the same fraction. One
aluminium measurement can serve all seven targets taken from the same sample, so
the second column is not a subset of the first and can be larger.

```{r}
#| output: asis
al <- s_ex[s_ex$symbol == "AL", ]
cat(sprintf("%s of what this site reports from this archive has an aluminium measured beside it, which is the ceiling on how much of it could ever carry an enrichment factor. %s See [How a pristine sample is decided](methods-pristine.qmd).\n\n",
            sprintf("%.1f%%", al$pct_paired),
            if (al$pct_paired >= 60)
              "Where one is actually published is a narrower question again, and depends on the element and on the fraction."
            else "Nothing later in the pipeline can lift it: without an aluminium on the same sample and the same fraction there is nothing to normalise against."))

absent <- s_ex[!s_ex$in_archive, ]
if (nrow(absent))
  cat(sprintf("This archive reports no %s at all.\n",
              tolower(and_list(absent$name))))
```

## How much survives each step

Counted over the seven target elements only. The normalisers, organic carbon and
the grain-size parameters travel through the same pipeline and are not counted
here, because this table is about the measurements the study is built on.

```{r}
s_fl |>
  transmute(Step = stage_label[stage],
            Measurements = format(n, big.mark = " "),
            Sites = ifelse(is.na(n_sites), "", format(n_sites, big.mark = " ")),
            `Of previous` = ifelse(is.na(pct_of_prev), "",
                                   sprintf("%.1f%%", pct_of_prev)),
            `Of parse` = sprintf("%.1f%%", pct_of_pilot),
            `What happens` = what) |>
  kable(align = c("l", "r", "r", "r", "r", "l"))
```

```{r}
#| fig-height: 2.9
#| fig-cap: !expr sprintf("Target-element measurements surviving each step, %s.", s_lab)
s_fl |>
  mutate(Step = factor(stage_label[stage], levels = unname(stage_label))) |>
  ggplot(aes(Step, n)) +
  geom_col(fill = unname(src_palette[s_lab]), width = .68) +
  geom_text(aes(label = format(n, big.mark = " ")), vjust = -0.4, size = 3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = NULL, y = "Measurements") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank(),
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r}
#| output: asis
# The parse-to-harmonise rung is the one a reader is most likely to misread, so
# it is stated rather than left to the percentage. Restricted to the targets it
# is all but lossless everywhere; over the whole archive it would not be, because
# the pilot database holds every parameter the source publishes.
lost_parse <- n_pilot - n_slim
cat(if (lost_parse == 0)
      "Nothing is lost between the parse and the harmonised schema: every target measurement the archive publishes reaches the shared tables.\n\n"
    else sprintf("Between the parse and the harmonised schema %s of %s measurements do not carry across: a small residue the parse could not resolve into the shared tables.\n\n",
                 format(lost_parse, big.mark = " "),
                 format(n_pilot, big.mark = " ")))

merge_pct <- s_fl$pct_of_prev[s_fl$stage == "merged"]
if (!is.na(merge_pct) && merge_pct < 99) {
  cat(sprintf("The merge step costs this archive %.1f%% of what reached it. That is where a cruise reported into two archives is reduced to one copy, and the copy kept is the one from the archive closest to the measurement. Nothing is lost from the study by it: those rows are still reported, under the other source's name.\n",
              100 - merge_pct))
} else {
  cat("The merge step costs this archive almost nothing, so very little of what it reports was also reported by one of the other four.\n")
}
```

## Why the rows that leave at the cleaning step leave

Each measurement is counted once, against the first rule that applied to it, so
these add up rather than overlap.

```{r}
s_rm |>
  filter(!is.na(n), n > 0) |>
  transmute(Reason = removal_label[reason],
            Measurements = format(n, big.mark = " "),
            `Of harmonised` = sprintf("%.1f%%", pct_of_slim),
            `What it means` = what) |>
  kable(align = c("l", "r", "r", "l"))
```

```{r}
#| output: asis
if (any(!is.na(s_rm$n) & s_rm$n == 0))
  cat("The reasons not listed removed nothing from this archive.\n\n")
comb <- s_rm$n[s_rm$reason == "combined"]
rej  <- sum(s_rm$n[s_rm$reason != "combined"], na.rm = TRUE)
if (!is.na(comb) && comb > rej)
  cat("Most of what leaves here is not a rejection. The archive reports the same sample more than once for one occasion and one method, and those repeats are averaged into a single row rather than discarded: the measurements are still in the number above, they are no longer separate rows.\n")
```

## Where its samples are

```{r}
#| include: false
sm <- read.csv(file.path(sdir, "summary_source_map.csv"), stringsAsFactors = FALSE)
sm <- sm[sm$key == SRC, c("lat", "lon", "n_sites", "n", "n_elements",
                          "year_min", "year_max", "depth_p50",
                          "n_bulk", "n_sieved63", "n_sieved20", "frac_top")]
ojs_define(onemap = sm)
# The fraction key, so the map can be coloured by what was sampled as well as by
# how much of it. Only the fractions this archive actually worked in are passed,
# which is also what the legend then lists.
fr_seen <- intersect(cat_order, unique(sm$frac_top))
ojs_define(fraclv = data.frame(cat = fr_seen,
                               label = unname(frac_label[fr_seen]),
                               colour = unname(frac_palette[fr_seen]),
                               stringsAsFactors = FALSE))
lat_lo <- min(sm$lat); lat_hi <- max(sm$lat)
depth_mid <- stats::median(sm$depth_p50, na.rm = TRUE)
```

One cell is 0.1 degrees, roughly 11 km, the same grid the element maps use.
Colour is sampling effort, or the fraction that was sampled, and never a
concentration: this map answers where and how this archive worked, not what was
found there. Its cells run from
`r sprintf("%.1f", lat_lo)` to `r sprintf("%.1f", lat_hi)` degrees north, and the
middle cell sits at
`r if (is.na(depth_mid)) "an unrecorded depth" else paste0(round(depth_mid), " m")`.

```{=html}
<link rel="stylesheet" href="https://unpkg.com/leaflet@1.9.4/dist/leaflet.css" />
```

```{ojs}
//| echo: false
L = require("https://unpkg.com/leaflet@1.9.4/dist/leaflet.js")
cells = transpose(onemap)
fracs = transpose(fraclv)
```

```{ojs}
//| echo: false
// Four ways of asking how much work happened in a cell, and one asking what kind
// of work it was. None of them is a concentration, which is the point: the
// element pages own those scales.
// The fraction option is offered only where there is more than one fraction to
// show: an archive that only ever sampled bulk would get a one-colour map and a
// control that does nothing.
layer_opts = [["Sites", "n_sites"], ["Measurements", "n"],
              ["Elements measured", "n_elements"], ["Median depth (m)", "depth_p50"]]
  .concat(fracs.length > 1 ? [["Fraction sampled", "frac_top"]] : [])
viewof layer = Inputs.radio(new Map(layer_opts),
                            { value: "n_sites", label: "Colour by" })
layer_name = (layer_opts.find(o => o[1] === layer) ?? ["", ""])[0]
```

```{ojs}
//| echo: false
srcMapDiv = {
  const div = html`<div id="map"></div>`;
  yield div;
  const map = L.map(div, { zoomControl: true });
  L.tileLayer("https://tile.openstreetmap.org/{z}/{x}/{y}.png", {
    attribution: "&copy; OpenStreetMap contributors",
    maxZoom: 12
  }).addTo(map);

  // The home view is this archive's own extent. The zoom floor of 3 is the one
  // the element maps use: Mercator stretches the far north, so fitting a range
  // that reaches Svalbard honestly would open on a world map.
  const lats = cells.map(d => d.lat), lons = cells.map(d => d.lon);
  const home = cells.length
    ? L.latLngBounds([Math.min(...lats), Math.min(...lons)],
                     [Math.max(...lats), Math.max(...lons)])
    : L.latLngBounds([36, -25], [81, 40]);
  const goHome = () => {
    map.invalidateSize();
    map.fitBounds(home, { padding: [24, 24], maxZoom: 7 });
    if (map.getZoom() < 3) map.setView(home.getCenter(), 3);
  };
  goHome();

  const HomeControl = L.Control.extend({
    options: { position: "topleft" },
    onAdd() {
      const btn = L.DomUtil.create("a", "leaflet-home-button");
      btn.href = "#";
      btn.title = "Reset the view";
      btn.setAttribute("aria-label", "Reset the view");
      btn.innerHTML = '<i class="bi bi-house-door" aria-hidden="true"></i>';
      L.DomEvent.on(btn, "click", e => { L.DomEvent.stop(e); goHome(); });
      return btn;
    }
  });
  new HomeControl().addTo(map);

  // The fraction layer is a category rather than an amount, so it takes its own
  // colours and its own legend. Everything else is a count and shares one ramp.
  const PLAIN = "#9e9e9e";
  const catg = layer === "frac_top";
  const fracColour = new Map(fracs.map(o => [o.cat, o.colour]));
  const fracLabel = new Map(fracs.map(o => [o.cat, o.label]));

  // Scaled on this archive's own cells, and clipped at the 5th and 95th
  // percentile so one dense cell does not flatten the rest. The legend says so.
  const vals = catg ? []
    : cells.map(d => d[layer]).filter(v => v !== null && !isNaN(v));
  const sorted = vals.slice().sort((a, b) => a - b);
  const q = p => sorted[Math.floor(p * (sorted.length - 1))];
  const lo = q(0.05), hi = q(0.95);
  // A measure this archive never recorded (depth, for one that reports none)
  // leaves nothing to scale. The cells are still places it sampled, so they are
  // drawn flat rather than dropped, and the legend says what is missing.
  const scaled = vals.length > 0;
  const ramp = scaled
    ? d3.scaleSequential(d3.interpolateViridis).domain([lo, hi]) : null;
  const colourOf = d => catg ? (fracColour.get(d.frac_top) ?? PLAIN)
                             : (scaled ? ramp(d[layer]) : PLAIN);
  // What a cell holds in each fraction, in every popup: the colour of a mixed
  // cell is the fraction with the most measurements in it, and this is what that
  // call was made on.
  const fracline = d => [["n_bulk", "bulk"], ["n_sieved63", "sieved <63 um"],
                         ["n_sieved20", "sieved <20 um"]]
    .filter(([k]) => d[k] > 0).map(([k, l]) => `${l} ${d[k]}`).join(", ");

  const drawn = cells.filter(d => catg || (d[layer] !== null && !isNaN(d[layer])));
  // biggest last, so a cell worth seeing is not painted over by a quiet one
  drawn.sort((a, b) => catg ? a.n - b.n : a[layer] - b[layer]);
  for (const d of drawn) {
    L.circleMarker([d.lat, d.lon], {
      radius: 4, weight: 0.4, color: "#33333355",
      fillColor: colourOf(d), fillOpacity: 0.85
    }).bindPopup(
      `<b>${d.n_sites} site${d.n_sites === 1 ? "" : "s"}</b>` +
      `<br/>${d.n} measurements, ${d.n_elements} element${d.n_elements === 1 ? "" : "s"}` +
      `<br/>${d.year_min} to ${d.year_max}` +
      (d.depth_p50 === null || isNaN(d.depth_p50) ? "" : `<br/>median depth ${d.depth_p50} m`) +
      `<br/>${fracline(d)}`
    ).addTo(map);
  }
  // cells the chosen measure says nothing about are still places this archive
  // sampled, so they stay on the map in one flat colour rather than vanishing.
  // The fraction layer has no such cells: every cell was sampled in some
  // fraction, and isNaN would call every one of its labels missing.
  for (const d of (catg ? [] : cells.filter(d => d[layer] === null || isNaN(d[layer])))) {
    L.circleMarker([d.lat, d.lon], {
      radius: 2.5, weight: 0, fillColor: PLAIN, fillOpacity: 0.3
    }).bindPopup(`<b>${d.n_sites} site${d.n_sites === 1 ? "" : "s"}</b><br/>No ${layer_name.toLowerCase()} recorded`)
      .addTo(map);
  }

  const fmt = v => Number.isInteger(v) ? v : v.toFixed(1);
  const n_grey = cells.length - drawn.length;
  const legend = L.control({ position: "bottomright" });
  legend.onAdd = () => {
    const box = L.DomUtil.create("div", "legend");
    const bar = scaled ? d3.range(0, 1.001, 0.05)
      .map(t => ramp(lo + t * (hi - lo))).join(", ") : "";
    if (catg) {
      box.innerHTML =
        `<div class="legend-title">${layer_name}</div>` +
        `<div class="legend-sub">per 0.1 degree cell</div>` +
        fracs.map(o =>
          `<div class="legend-row"><span class="legend-swatch" ` +
          `style="background:${o.colour}"></span>${o.label}</div>`).join("") +
        `<div class="legend-note">A cell takes the colour of the fraction it holds ` +
        `most measurements in. The popup gives the counts.</div>`;
      return box;
    }
    box.innerHTML =
      `<div class="legend-title">${layer_name}</div>` +
      `<div class="legend-sub">per 0.1 degree cell</div>` +
      (scaled
        ? `<div class="legend-barwrap"><div class="legend-bar" ` +
          `style="background: linear-gradient(to right, ${bar})"></div></div>` +
          `<div class="legend-labels"><span>${fmt(lo)} and below</span>` +
          `<span>${fmt(hi)} and above</span></div>`
        : `<div class="legend-row"><span class="legend-swatch" ` +
          `style="background:${PLAIN}"></span>Cells with data</div>`) +
      (n_grey ? `<div class="legend-note">${n_grey} cell${n_grey === 1 ? "" : "s"} ` +
                `with nothing recorded, drawn faint.</div>` : "") +
      (scaled ? "" : `<div class="legend-note">This archive records no ` +
                     `${layer_name.toLowerCase()}, so nothing is scaled here.</div>`);
    return box;
  };
  legend.addTo(map);
}
```

A cell is a summary of everything this archive did inside it, so a cell holding
one deep station and one shallow one reports the middle of them. The per-site
detail is in `summary_source_map.csv`'s full-resolution companion,
`summary_map_sites.csv`, on the [Downloads](downloads.qmd) page.
