---
title: "Downloads"
subtitle: "The database, the flat dataset, and the tables behind this site"
---

{{< include _setup.qmd >}}

Everything here is published from the refined generation, which is the database
this site summarises. There is deliberately no second copy of the data hosted on
this site: one published copy cannot go stale against another.

## The analysis database

```{r}
#| output: asis
keyfig(
  list(format(smeta$n_measurement, big.mark = "\u00a0"), "measurements"),
  list(format(smeta$n_sites, big.mark = "\u00a0"), "sites"),
  list(smeta$n_elements, "target elements"),
  list(smeta$db_modified, "database built")
)
```

**[multised_refined.sqlite](https://github.com/seafood-hazards/multised-refined/releases/latest/download/multised_refined.sqlite)**
(SQLite, about 22 MB)

The full relational database: measurements, sites, sampling events, methods,
normalisers and the aquaculture register, with every derived flag this site
reports. Use this if you want to run your own queries.

Schema and table documentation:
[DB Schema](`r REFINED`/db-schema-refined.html).

## The flat dataset

**[multised_refined_dataset.tsv.gz](https://github.com/seafood-hazards/multised-refined/releases/latest/download/multised_refined_dataset.tsv.gz)**
(gzipped TSV, about 3 MB)

One row per measurement with every joined column already attached, plus a column
dictionary. Use this if you want to open the data in R, Python, or a spreadsheet
without writing SQL.

Column-by-column description:
[Dataset Download](`r REFINED`/download-dataset.html).

## The EFSA submission table

**[EFSA format](`r REFINED`/efsa-format-v2.html)** on the detailed site.

A separate cut of the same data, shaped to EFSA's reporting workbook and its
extraction specification. Both it and the flat dataset are cut from one frame, so
they cannot disagree with each other.

::: {.callout-note}
## Not the final submission

The dataset actually submitted will be a filtered subset of this, restricted to
samples with a pristine verdict. That filtering step is not built yet, so no
submission file is published here.
:::

## The tables behind this site

Every figure and table on this site is read from these files. They are attached to
each [release of this repository](https://github.com/seafood-hazards/multised-summary/releases/latest).

```{r}
# The descriptions are written here, but the file list is checked against
# _scripts/release-assets.txt so this page cannot quietly fall behind a release
# that carries a file it does not describe.
tbl <- data.frame(
  File = c("summary_elements.csv", "summary_reference.csv", "summary_sea_spread.csv",
           "summary_background.csv", "summary_verdicts.csv",
           "summary_coverage.csv", "summary_censoring.csv", "summary_pressure.csv",
           "summary_gradient.csv", "summary_controls.csv", "summary_sources.csv",
           "summary_flow.csv", "summary_map_grid.csv", "summary_map_sites.csv",
           "summary_meta.csv"),
  Contents = c(
    "One row per element and fraction: how much data, and what can be said about it",
    "What each fraction's offshore reference population actually is: where it sits, how deep, and how little of it lies in the far north",
    "How far each background moves between seas, with and without aluminium: which grain-size control travels",
    "Every background estimate from every method",
    "Classifiable share and pristine share, per element and fraction",
    "How the classifiable share falls with distance to a fish farm and to the coast",
    "Below-detection-limit shares per element and source",
    "The distance-band gradient, as concentration and as Igeo",
    "Near-versus-far enrichment, before and after removing monitored pollution",
    "The four independent controls on the near-cage gradient",
    "What each of the five sources contributes",
    "Measurement counts at each pipeline stage",
    "A 0.1 degree grid of median concentration and Igeo: the map layer",
    "The same at full site resolution",
    "Provenance: which database, built when, with which rules"
  )
)

manifest <- readLines("_scripts/release-assets.txt")
manifest <- trimws(manifest)
manifest <- manifest[nzchar(manifest) & !startsWith(manifest, "#")]
undescribed <- setdiff(manifest, tbl$File)
if (length(undescribed))
  stop("release assets with no description on this page: ",
       paste(undescribed, collapse = ", "))

kable(tbl)
```

## Licence and citation

The underlying measurements belong to their original providers
(`r paste(sources$source, collapse = ", ")`) and remain subject to their terms.
The harmonised database, the derived values and this site are produced by the
Institute of Marine Research.

```{r}
#| output: asis
cat("Built from `", smeta$refined_db, "` (", smeta$db_modified, "), assembled ",
    smeta$generated, ".\n", sep = "")
```
