---
title: "Background and pristine marine sediment"
subtitle: "Seven trace elements, five data sources, one summary"
toc: false
---

{{< include _setup.qmd >}}

We collected every marine-sediment trace-element measurement we could obtain from
five public sources, harmonised them into one database, and asked two questions of
it:

1. **What is the natural background level of each element?**
2. **Which samples are at that background, i.e. pristine?**

This site is the short answer. Every number on it is computed by the pipeline and
read from a file; nothing is typed in by hand. Where you want the workings, each
page links to the corresponding page on the
[detailed analysis site](`r REFINED`).

```{r}
#| output: asis
tot <- flow$n[flow$stage == "analysed"]
keyfig(
  list(format(tot, big.mark = " "), "measurements"),
  list(format(smeta$n_sites, big.mark = " "), "sampling sites"),
  list(nrow(sources), "data sources"),
  list(paste0(min(sources$year_min), " to ", max(sources$year_max)), "years covered"),
  list("3 of 7", "elements with a pristine verdict")
)
```

## How the data was built

```{r}
#| output: asis
#| fig-cap: "Five sources become one database."
# The counts are read from summary_flow.csv, not typed here: the diagram cannot
# drift away from the databases it describes.
n <- function(stage) format(flow$n[flow$stage == stage], big.mark = " ")
src <- paste0(sprintf('  %s["%s"] --> S\n', LETTERS[seq_len(nrow(sources))],
                      sources$source), collapse = "")
cat("```{mermaid}\n",
    "flowchart LR\n", src,
    sprintf('  S["Harmonised<br/>%s"] --> Cl["Quality controlled<br/>%s"]\n', n("slim"), n("clean")),
    sprintf('  Cl --> M["Merged, de-duplicated<br/>%s"]\n', n("merged")),
    sprintf('  M --> R["Analysis database<br/>%s"]\n', n("refined")),
    sprintf('  R --> A2["Reported here<br/>%s"]\n', n("analysed")),
    '  A2 --> Q1["Background"]\n',
    '  A2 --> Q2["Pristine verdict"]\n',
    "  style A2 fill:#e8f0fe,stroke:#4477AA\n",
    "  style Q1 fill:#d7ecd9,stroke:#1c5626\n",
    "  style Q2 fill:#d7ecd9,stroke:#1c5626\n",
    "```\n", sep = "")
```

Details: [Data Collection and Preparation](methods-data.qmd).

## What we can say, element by element

Not every element can be reported the same way, and the reason is always in the
data rather than in the method. Two rules decide it, and both are applied before
any result is looked at:

- An element whose measurements were mostly **below the limit of quantification**
  has no usable background. What survives is an upper tail wearing the name
  background. This removes selenium and molybdenum.
- A **pristine verdict** needs aluminium to control for grain size, and aluminium
  only predicts an element for some of them. Where it does not, dividing by it is
  not a grain-size control, so we publish concentrations and background but issue
  no verdict.

```{r}
#| output: asis
tab <- elements |>
  group_by(symbol, name) |>
  summarise(n = first(n_elem),
            cap = capability(first(symbol)),
            .groups = "drop") |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  arrange(symbol) |>
  transmute(Element = sprintf("[%s (%s)](element-%s.qmd)", name,
                              sym_display[as.character(symbol)],
                              tolower(as.character(symbol))),
            Measurements = format(n, big.mark = " "),
            `What is published` = chip(cap),
            Why = vapply(as.character(symbol), capability_note, character(1)))
kable(tab, escape = FALSE)
```

## The aquaculture question

Norway's fish farms are a plausible source of copper and zinc to the seabed. We
tested it on an axis the background work never saw, so the two are independent.

```{r}
#| output: asis
g <- gradient |> filter(cat == "bulk", symbol %in% c("CU", "ZN", "CO")) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |> arrange(symbol)
cu <- g$enrich_near_clean[g$symbol == "CU"]
n_ctl <- sum(controls$symbol == "CU" & controls$cat == "bulk" & controls$reliable)
keyfig(
  list(sprintf("%.1fx", cu), "copper under the cages vs beyond 20 km"),
  list(sprintf("%d of %d", n_ctl, n_ctl), "independent controls copper survives"),
  list("no", "verdict changes: the gradient is measured, not assumed")
)
```

Details: [Aquaculture Pressure](methods-pressure.qmd).

::: {.callout-note}
## This is not the final submission

This site summarises the analysis as it stands. The dataset submitted to EFSA is
a filtered subset of it, and that filtering step is not built yet. Treat the
numbers here as current, not final.
:::
