---
title: "Estimation of Pristine Sediment"
subtitle: "Deciding whether one sample is at its background, and admitting when we cannot"
---

{{< include _setup.qmd >}}

A background is a number for a region. A **pristine verdict** is a judgement about
a single sample: is this one at its natural level, or above it?

The obstacle is grain size. Fine, muddy sediment naturally holds several times more
metal than coarse sand from the same seabed, with no pollution involved. Comparing
a raw concentration against a regional background therefore measures texture as
much as contamination.

## Two ways round it

There are two, and which one applies is decided by the sample, not by us: either
the grain size was removed in the laboratory by a sieve, or it has to be corrected
for arithmetically, by aluminium.

```{r}
#| output: asis
cat("```{mermaid}\n",
    "flowchart TD\n",
    '  S["One sample"] --> Q{"Was it sieved<br/>before analysis?"}\n',
    '  Q -->|yes| SV["<b>The sieve</b> is the<br/>grain-size control"]\n',
    '  Q -->|"no, bulk sediment"| A{"Does aluminium predict<br/>this element, and was<br/>it measured?"}\n',
    '  A -->|yes| EF["<b>Enrichment factor</b><br/>metal per aluminium,<br/>against an offshore reference"]\n',
    '  A -->|no| N["No verdict:<br/>cannot be classified"]\n',
    '  EF --> V["EF below 1<br/>= pristine"]\n',
    '  SV --> W["below both concentration<br/>criteria = pristine,<br/>strict rule only"]\n',
    "  style EF fill:#e8f0fe,stroke:#4477AA\n",
    "  style SV fill:#e8f0fe,stroke:#4477AA\n",
    "  style N fill:#eeeeee,stroke:#888888\n",
    "```\n", sep = "")
```

**Aluminium is the grain-size correction.** It is a structural component of clay,
so it rises and falls with the mud content of a sample. Dividing the metal by it
removes the texture effect and leaves what is left over.

::: {.callout-note}
## The geo-accumulation index is reported, not used to judge

Igeo compares a sample with a regional background concentration for the same
element. It needs no aluminium, so it reaches almost every measurement we hold,
where the enrichment factor reaches a small minority of them, and EFSA names it as
an alternative where no enrichment factor is available.

It is still not what decides anything here. Igeo divides by a background, not by a
normaliser, so it does not correct for grain size at all: in bulk sediment it
tracks the mud content of a sample closely enough that a verdict built on it would
be partly a verdict about texture. A muddy natural seabed would read as polluted.

So Igeo appears throughout this site, on the element pages, on the maps and in the
downloadable data, as a second reading to set beside the verdict. **No pristine
verdict uses it, and it plays no part in what is submitted.** The
[Geo-Accumulation Index](`r REFINED`/background-igeo.html) page on the refined site
has the measurements behind that decision.
:::

::: {.callout-warning}
## Iron is deliberately not used

Iron is the other conventional normaliser and would give far better coverage.
Norwegian fish feed contains iron, so iron near a farm is partly the very pressure
being measured. Using it would quietly subtract the signal. Iron is reported for
comparison and never used to issue a verdict.
:::

## In bulk sediment, aluminium only works for three of the seven

A normaliser is only a grain-size control if it actually predicts the element.
Measured against the offshore reference, aluminium predicts cobalt, copper and zinc
in bulk sediment, and nothing else.

```{r}
elements |>
  filter(cat == "bulk") |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  arrange(symbol) |>
  transmute(Element = paste0(name, " (", sym_display[as.character(symbol)], ")"),
            `Aluminium predicts it in bulk` = ifelse(normalisable, "yes", "no"),
            `Bulk verdict issued` = ifelse(has_verdict, "yes", "no"),
            `What is published overall` =
              vapply(as.character(symbol), capability_note, character(1))) |>
  kable()
```

Where aluminium fails, we issue no enrichment factor. The alternative would be to
divide by something that does not track texture and call the result a grain-size
control, which would be worse than saying nothing.

The last column looks past bulk sediment, and for manganese the two disagree: it
gets no bulk verdict and does get a sieved one. The next section is why.

## The sieved samples are judged on the sieve

Some of our samples were sieved before analysis, cut to particles below 63 or 20
micrometres. Sieving is itself a grain-size correction, so aluminium has nothing
left to track in a sieved sample and duly fails the test above.

**That is the sieve working, not aluminium failing**, and for a while this site
read it the wrong way round. Until August 2026 a sieved sample got no verdict at
all, on the grounds that aluminium did not predict the metal in it. That treated
"no correction needed" as "the correction failed", and penalised the samples that
had the better grain-size control to begin with.

Better is measurable. A grain-size control that works should leave a background
close to the same number wherever you look. Taking each group's offshore
background sea by sea, and dividing the highest sea by the lowest:

```{r}
sea_spread |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         Fraction = factor(cat_label[cat], levels = cat_label[cat_order])) |>
  arrange(symbol, Fraction) |>
  transmute(Element = sym_display[as.character(symbol)], Fraction,
            `Seas compared` = n_seas,
            `Concentration alone` = paste0(fold_raw, " times"),
            `After dividing by aluminium` = paste0(fold_al, " times")) |>
  kable()
```

The sieved rows come closest to holding one number across the map, and they do it
in the first column, before aluminium is applied at all. So the sieved samples are
judged without an enrichment factor, on the two concentration criteria described
in the next section. Aluminium is still measured and still reported for them. It
just does not decide anything.

### What the sieved verdicts still cannot tell you

One caveat travels with them, and it is not a small one. Sieving is a national
convention rather than a property of a sample: some countries sieve and report
sieved values, others report the bulk sediment. So "offshore" is not the same
place in the two fractions.

```{r}
reference |>
  mutate(Fraction = factor(cat_label[cat], levels = cat_label[cat_order])) |>
  arrange(Fraction) |>
  transmute(Fraction, `Grain-size control` = label,
            `Reference samples` = format(n, big.mark = "\u00a0"),
            `Typical latitude` = paste0(lat_p50, " degrees north"),
            `Typical depth` = paste0(depth_p50, " m"),
            `Mostly` = top_seas,
            `Far north` = format(n_north, big.mark = "\u00a0")) |>
  kable(col.names = c("Fraction", "Grain-size control", "Reference samples",
                      "Typical latitude", "Typical depth", "Mostly",
                      sprintf("Above %g degrees north", reference$north_cut_deg[1])))
```

The bulk background is the deep Norwegian and Barents Sea. The sieved background
is the shallow southern North Sea and Baltic, off an industrialised continental
coast. The last column is why this cannot simply be fixed by re-cutting the
reference: there is almost no sieved data in the far north to build a northern
background from.

So a sieved verdict is an honest answer to "is this sample at or below background
**for its own fraction**", and it is not evidence about Norwegian waters. That is
also why the sieved samples get the strict verdict only, and never the permissive
one.

## The verdict, and its cost

Two rules are published. The permissive one is a single test, the enrichment
factor. The strict one asks every criterion that applies to the fraction to agree:
in bulk that is the enrichment factor plus two concentration criteria, and in the
sieved fractions it is those two concentration criteria, standing on the sieve for
their grain-size control.

Bulk sediment first.

```{r}
verdicts |>
  filter(cat == "bulk", reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  arrange(symbol) |>
  transmute(Element = sym_display[as.character(symbol)],
            Measurements = format(n, big.mark = "\u00a0"),
            `Can be judged` = sprintf("%s%%", pct_classifiable),
            `Judged pristine` = sprintf("%s%%", pct_ef),
            `Pristine, strict rule` = sprintf("%s%%", pct_strict)) |>
  kable()
```

The middle column is the finding that matters most, and it is a data gap rather
than a result. **Most bulk samples cannot be judged at all**, because aluminium was
never measured alongside the metal.

The sieved samples need no aluminium, so almost none of them are lost this way.
They carry the strict rule only, for the reason given above: their background is a
different sea.

```{r}
verdicts |>
  filter(cat != "bulk", reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         Fraction = factor(cat_label[cat], levels = cat_label[cat_order])) |>
  arrange(symbol, Fraction) |>
  transmute(Element = sym_display[as.character(symbol)], Fraction,
            Measurements = format(n, big.mark = "\u00a0"),
            `Can be judged` = sprintf("%s%%", pct_classifiable),
            `Pristine, strict rule` = sprintf("%s%%", pct_strict)) |>
  kable()
```

The gap that follows is a gap in the aluminium, so it is a bulk problem only.

```{r}
#| fig-height: 2.8
#| fig-cap: "The share of samples that can be judged falls away as you approach a fish farm."
cov <- coverage |>
  filter(axis == "distance to fish farm") |>
  mutate(band = factor(band, levels = band))
ggplot(cov, aes(band, pct_classifiable)) +
  geom_col(fill = "#CC6677") +
  geom_text(aes(label = paste0(pct_classifiable, "%")), vjust = -0.4, size = 3.3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = "Distance to the nearest fish farm", y = "Can be judged (%)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())
```

Aluminium coverage runs opposite to what the question needs: it is common in
offshore survey work and rare in near-cage monitoring. Almost none of the sediment
under a Norwegian fish farm can be given an aluminium-based verdict, which is why
the [aquaculture question](methods-pressure.qmd) is answered with a different tool.

Full workings: [Enrichment Factor](`r REFINED`/background-ef.html),
[Pristine Classification](`r REFINED`/background-pristine.html),
[Regression Normalisation](`r REFINED`/background-regression.html).
