---
title: "Data Collection and Preparation"
subtitle: "How five separate archives became one comparable database"
---

{{< include _setup.qmd >}}

## The five sources

Each source is a public archive with its own file formats, units, element codes and
quality conventions. None of them was designed to be combined with the others.

```{r}
sources |>
  transmute(Source = source,
            Measurements = format(n, big.mark = "\u00a0"),
            `Share` = sprintf("%.1f%%", pct),
            Sites = format(n_sites, big.mark = "\u00a0"),
            Datasets = n_datasets,
            Elements = n_elements,
            Years = paste0(year_min, " to ", year_max)) |>
  kable()
```

## The five steps

Each step is a separate database, so any result can be traced back and any step
re-run on its own.

```{r}
#| output: asis
cat("```{mermaid}\n",
    "flowchart TD\n",
    '  P["1. Parse<br/><i>read each archive as it is</i>"]\n',
    '  S["2. Harmonise<br/><i>one 7-table schema, shared units</i>"]\n',
    '  C["3. Clean<br/><i>apply the quality flags</i>"]\n',
    '  M["4. Merge<br/><i>union, then remove duplicates</i>"]\n',
    '  R["5. Refine<br/><i>cut the analysis database</i>"]\n',
    "  P --> S --> C --> M --> R\n",
    "  style R fill:#e8f0fe,stroke:#4477AA\n",
    "```\n", sep = "")
```

```{r}
step_label <- c(slim = "2. Harmonise", clean = "3. Clean", merged = "4. Merge",
                refined = "5. Refine", analysed = "Reported on this site")
flow |>
  transmute(Step = step_label[stage],
            Measurements = format(n, big.mark = "\u00a0"),
            Sites = format(n_sites, big.mark = "\u00a0"),
            `What happens` = what) |>
  kable()
```

```{r}
#| include: false
drop_pct <- round(100 * (1 - flow$n[flow$stage == "analysed"] /
                             flow$n[flow$stage == "slim"]))
```

About `r drop_pct` per cent of the harmonised measurements do not reach the final
reported set. That is the point of the pipeline rather than a loss: the rows
removed are failed quality checks, values below the limit of quantification,
elements outside this study, and the same sample reported twice by two archives.

## What had to be made comparable

```{r}
tibble::tribble(
  ~Problem, ~`What we did`,
  "Different units (mg/kg, ug/g, ppm, %)", "Converted everything to mg/kg dry weight",
  "Different element codes per archive",   "Mapped to one element table with CAS numbers",
  "Wet-weight and dry-weight reporting",   "Recorded the basis, and standardised to dry weight",
  "Bulk sediment vs sieved fractions",     "Kept them apart: they are different materials, never pooled",
  "The same cruise in two archives",       "Matched on position, date and value, and kept one",
  "Positions of varying precision",        "Sites keyed on latitude and longitude to 3 decimal places"
) |> kable()
```

::: {.callout-important}
## Bulk and sieved samples are never mixed

Sieving concentrates fine particles, and fine particles carry more metal. A sieved
sample and a bulk sample from the same seabed give different numbers for the same
sediment. Every result on this site is reported separately for bulk, for the
fraction below 63 um and for the fraction below 20 um, and no average is ever taken
across them.

One consequence is worth knowing early: because the two fractions have their grain
size controlled in different ways, they are judged by different criteria. Bulk
sediment is corrected using aluminium; a sieved sample was corrected by the sieve
before it was analysed, so it is judged on concentration alone. Both can carry a
pristine verdict, and for manganese only the sieved one does. See
[Estimation of Pristine Sediment](methods-pristine.qmd#the-sieved-samples-are-judged-on-the-sieve).
:::

Full detail: [database schema](`r REFINED`/db-schema-refined.html) and the
[cleaning report](https://seafood-hazards.github.io/multised-clean) on the
detailed sites.
