---
title: "Estimation of Pristine Sediment"
subtitle: "Deciding whether one sample is at its background, and admitting when we cannot"
---

{{< include _setup.qmd >}}

A background is a number for a region. A **pristine verdict** is a judgement about
a single sample: is this one at its natural level, or above it?

The obstacle is grain size. Fine, muddy sediment naturally holds several times more
metal than coarse sand from the same seabed, with no pollution involved. Comparing
a raw concentration against a regional background therefore measures texture as
much as contamination.

## Two ways round it

```{r}
#| output: asis
cat("```{mermaid}\n",
    "flowchart TD\n",
    '  S["One sample"] --> Q{"Is aluminium<br/>measured?"}\n',
    '  Q -->|yes| EF["<b>Enrichment factor</b><br/>metal per aluminium,<br/>against an offshore reference"]\n',
    '  Q -->|no| I["<b>Geo-accumulation index</b><br/>metal against a regional<br/>background concentration"]\n',
    '  EF --> V["EF below 1<br/>= pristine"]\n',
    '  I --> W["Igeo at or below 0<br/>= unpolluted"]\n',
    "  style EF fill:#e8f0fe,stroke:#4477AA\n",
    "  style I fill:#fdf0d0,stroke:#7a5300\n",
    "```\n", sep = "")
```

**Aluminium is the grain-size correction.** It is a structural component of clay,
so it rises and falls with the mud content of a sample. Dividing the metal by it
removes the texture effect and leaves what is left over.

::: {.callout-warning}
## Iron is deliberately not used

Iron is the other conventional normaliser and would give far better coverage.
Norwegian fish feed contains iron, so iron near a farm is partly the very pressure
being measured. Using it would quietly subtract the signal. Iron is reported for
comparison and never used to issue a verdict.
:::

## Aluminium only works for three of the seven

A normaliser is only a grain-size control if it actually predicts the element.
Measured against the offshore reference, aluminium predicts cobalt, copper and zinc
in bulk sediment, and nothing else.

```{r}
elements |>
  filter(cat == "bulk") |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  arrange(symbol) |>
  transmute(Element = paste0(name, " (", sym_display[as.character(symbol)], ")"),
            `Aluminium predicts it` = ifelse(normalisable, "yes", "no"),
            `Verdict issued` = ifelse(has_verdict, "yes", "no"),
            Reason = vapply(as.character(symbol), capability_note, character(1))) |>
  kable()
```

Where aluminium fails, we issue no enrichment factor. The alternative would be to
divide by something that does not track texture and call the result a grain-size
control, which would be worse than saying nothing.

## And in bulk sediment only

Every result on this site is reported three times over: once for bulk sediment,
once for the fraction sieved below 63 um, once for the fraction below 20 um. The
pristine verdict is the exception, and exists for bulk alone.

It is the same test that decides it. **A sieved sample has already been grain-size
corrected, by the sieve.** Cut the sample to below 63 or 20 um and most of the
texture variation is gone before the chemistry starts, leaving aluminium almost
nothing to track. An enrichment factor built on it would be dividing by noise.

```{r}
elements |>
  mutate(Fraction = factor(cat_label[cat], levels = cat_label[cat_order])) |>
  group_by(Fraction) |>
  summarise(`Element groups` = n(),
            `Aluminium predicts the metal` = sum(normalisable),
            `Verdict issued` = sum(has_verdict),
            .groups = "drop") |>
  kable()
```

Sieved samples are not discarded. Their measurements, their background estimates
and their contamination classes are all published, the last of these because the
geo-accumulation index needs no aluminium at all. What a sieved sample does not get
is a pristine verdict. The maps on the element pages show bulk sediment for the
related reason: it holds by far the most data, and it is the only fraction the
verdict speaks about.

## The verdict, and its cost

Two rules are published. The permissive one is a single test; the strict one
requires three independent criteria to agree.

```{r}
verdicts |>
  filter(cat == "bulk", reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  arrange(symbol) |>
  transmute(Element = sym_display[as.character(symbol)],
            Measurements = format(n, big.mark = "\u00a0"),
            `Can be judged` = sprintf("%s%%", pct_classifiable),
            `Judged pristine` = sprintf("%s%%", pct_ef),
            `Pristine, strict rule` = sprintf("%s%%", pct_strict)) |>
  kable()
```

The middle column is the finding that matters most, and it is a data gap rather
than a result. **Most samples cannot be judged at all**, because aluminium was
never measured alongside the metal.

```{r}
#| fig-height: 2.8
#| fig-cap: "The share of samples that can be judged falls away as you approach a fish farm."
cov <- coverage |>
  filter(axis == "distance to fish farm") |>
  mutate(band = factor(band, levels = band))
ggplot(cov, aes(band, pct_classifiable)) +
  geom_col(fill = "#CC6677") +
  geom_text(aes(label = paste0(pct_classifiable, "%")), vjust = -0.4, size = 3.3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = "Distance to the nearest fish farm", y = "Can be judged (%)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())
```

Aluminium coverage runs opposite to what the question needs: it is common in
offshore survey work and rare in near-cage monitoring. Almost none of the sediment
under a Norwegian fish farm can be given an aluminium-based verdict, which is why
the [aquaculture question](methods-pressure.qmd) is answered with a different tool.

Full workings: [Enrichment Factor](`r REFINED`/background-ef.html),
[Pristine Classification](`r REFINED`/background-pristine.html),
[Regression Normalisation](`r REFINED`/background-regression.html).
