---
title: "Background Identification"
subtitle: "Five ways of asking what is natural, and what they agree on"
---

{{< include _setup.qmd >}}

The background is the concentration you would expect with no human input. Nobody
can measure it directly, because there is no untouched control coast to sample. So
it has to be inferred, and every way of inferring it rests on an assumption that
could be wrong.

We used five, deliberately chosen to fail in different ways.

```{r}
#| output: asis
cat("```{mermaid}\n",
    "flowchart LR\n",
    '  Q["What is<br/>natural?"]\n',
    '  Q --> A["<b>Where</b><br/>far from the coast"]\n',
    '  Q --> B["<b>How much</b><br/>the low percentiles"]\n',
    '  Q --> C["<b>Shape</b><br/>two overlapping populations"]\n',
    '  Q --> D["<b>Texture</b><br/>corrected for mud content"]\n',
    '  Q --> E["<b>Pressure</b><br/>far from any fish farm"]\n',
    "  style Q fill:#e8f0fe,stroke:#4477AA\n",
    "```\n", sep = "")
```

```{r}
tibble::tribble(
  ~Method, ~`The assumption`, ~`How it can fail`,
  "Offshore", "Sediment more than 10 km from the coast is little affected by land", "Offshore does not mean clean: some of it is monitored precisely because it is contaminated",
  "Percentiles", "The low end of the distribution is the natural end", "If most samples are polluted, the low end is polluted too",
  "Mixture", "Natural and enriched samples form two distinguishable populations", "Only works where the two are actually separable",
  "Grain-size normalised", "Compare like with like after correcting for mud content", "Needs the measured mud fraction, which most samples lack, and which describes the sediment as collected, so it applies to bulk only",
  "Pressure bands", "Sediment far from a known source is closer to natural", "Distance is a proxy: other sources exist"
) |> kable()
```

::: {.callout-tip}
## Agreement is the evidence

No single method is treated as correct. Where several agree, the agreement is the
result. Where they disagree, the disagreement is the result, and we say so instead
of picking a favourite.
:::

## What the five give

Bulk sediment, in `r UNIT`. A blank means the method could not be applied to that
element.

```{r}
# igeo_b is deliberately not shown: the Igeo reference IS the offshore median, so
# it would be a second identical column. It stays in summary_background.csv.
lvl <- c("offshore", "offshore_p90", "mixture", "gsnorm", "pressure_far")
lab <- c(offshore = "Offshore median", offshore_p90 = "Offshore P90",
         mixture = "Mixture bound", gsnorm = "Grain-size normalised",
         pressure_far = "Far from farms")
backgrounds |>
  filter(cat == "bulk", method %in% lvl) |>
  mutate(shown = ifelse(reliable, fmt(value), NA_character_),
         method = factor(lab[method], levels = lab[lvl]),
         symbol = factor(symbol, levels = elem_levels)) |>
  select(symbol, method, shown) |>
  arrange(symbol, method) |>
  pivot_wider(names_from = method, values_from = shown, names_sort = FALSE) |>
  mutate(symbol = sym_display[as.character(symbol)]) |>
  rename(Element = symbol) |>
  kable()
```

Two things stand out, and both are honest limits rather than problems to hide.

**Selenium and molybdenum are missing.** Over half their measurements fell below
what the laboratory could quantify, and those rows are removed during cleaning. A
background is the low end of a distribution, so removing the low end and then
estimating a background from what remains would give a number that is too high and
looks respectable. We publish no background for either.

```{r}
#| fig-height: 2.6
#| fig-cap: "Share of measurements below the limit of quantification. Above 20 per cent, the background is withheld."
elements |>
  group_by(symbol) |>
  summarise(pct = first(pct_censored), withheld = any(withheld), .groups = "drop") |>
  mutate(symbol = factor(sym_display[as.character(symbol)],
                         levels = rev(sym_display[elem_levels]))) |>
  ggplot(aes(pct, symbol, fill = withheld)) +
  geom_col() +
  geom_vline(xintercept = 20, linetype = 2, colour = "#b3261e") +
  geom_text(aes(label = sprintf("%.1f%%", pct)), hjust = -0.15, size = 3.1) +
  scale_fill_manual(values = c(`FALSE` = "#4477AA", `TRUE` = "#CC6677"), guide = "none") +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(x = "Below the limit of quantification (%)", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank())
```

**The methods disagree most for copper and zinc.** These are the two elements with
a real human source in this region, so a spread between "the low end of everything"
and "far from any farm" is what you would expect to see.

Full workings: [Background](`r REFINED`/background.html),
[Mixture](`r REFINED`/background-mixture.html),
[Grain-size normalised](`r REFINED`/background-gsnorm.html),
[Pressure-based](`r REFINED`/background-pressure.html).
