---
title: "Estimation of Pristine Sediment"
subtitle: "Deciding whether one sample is at its background, and admitting when we cannot"
---

{{< include _setup.qmd >}}

A background is a number for a region. A **pristine verdict** is a judgement about
a single sample: is this one at its natural level, or above it?

The obstacle is grain size. Fine, muddy sediment naturally holds several times more
metal than coarse sand from the same seabed, with no pollution involved. Comparing
a raw concentration against a regional background therefore measures texture as
much as contamination.

## Two ways round it

```{r}
#| output: asis
cat("```{mermaid}\n",
    "flowchart TD\n",
    '  S["One sample"] --> Q{"Is aluminium<br/>measured?"}\n',
    '  Q -->|yes| EF["<b>Enrichment factor</b><br/>metal per aluminium,<br/>against an offshore reference"]\n',
    '  Q -->|no| I["<b>Geo-accumulation index</b><br/>metal against a regional<br/>background concentration"]\n',
    '  EF --> V["EF below 1<br/>= pristine"]\n',
    '  I --> W["Igeo at or below 0<br/>= unpolluted"]\n',
    "  style EF fill:#e8f0fe,stroke:#4477AA\n",
    "  style I fill:#fdf0d0,stroke:#7a5300\n",
    "```\n", sep = "")
```

**Aluminium is the grain-size correction.** It is a structural component of clay,
so it rises and falls with the mud content of a sample. Dividing the metal by it
removes the texture effect and leaves what is left over.

::: {.callout-warning}
## Iron is deliberately not used

Iron is the other conventional normaliser and would give far better coverage.
Norwegian fish feed contains iron, so iron near a farm is partly the very pressure
being measured. Using it would quietly subtract the signal. Iron is reported for
comparison and never used to issue a verdict.
:::

## Aluminium only works for three of the seven

A normaliser is only a grain-size control if it actually predicts the element.
Measured against the offshore reference, aluminium predicts cobalt, copper and zinc
in bulk sediment, and nothing else.

```{r}
elements |>
  filter(cat == "bulk") |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  arrange(symbol) |>
  transmute(Element = paste0(name, " (", sym_display[as.character(symbol)], ")"),
            `Aluminium predicts it` = ifelse(normalisable, "yes", "no"),
            `Verdict issued` = ifelse(has_verdict, "yes", "no"),
            Reason = vapply(as.character(symbol), capability_note, character(1))) |>
  kable()
```

Where aluminium fails, we issue no enrichment factor. The alternative would be to
divide by something that does not track texture and call the result a grain-size
control, which would be worse than saying nothing.

## And in bulk sediment only

Every result on this site is reported three times over: once for bulk sediment,
once for the fraction sieved below 63 um, once for the fraction below 20 um. The
pristine verdict is the exception, and exists for bulk alone.

**It is not a shortage of sieved data.** That is the natural guess and it is wrong.
The sieved fractions carry thousands of measurements, and thousands of them have an
aluminium result alongside. The problem is what aluminium does in them.

```{r}
pct_expl <- function(r2) ifelse(is.na(r2), NA_character_,
                        ifelse(100 * r2 < 0.1, "under 0.1%",
                               sprintf("%.1f%%", 100 * r2)))
elements |>
  filter(symbol %in% c("CU", "ZN", "CO"), al_tested) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         Fraction = factor(cat_label[cat], levels = cat_label[cat_order])) |>
  arrange(symbol, Fraction) |>
  transmute(Element = sym_display[as.character(symbol)],
            Fraction,
            `Measurements` = format(n, big.mark = "\u00a0"),
            `Samples with aluminium` = format(al_n, big.mark = "\u00a0"),
            `Variation aluminium explains` = pct_expl(al_r2),
            `Verdict issued` = ifelse(has_verdict, "yes", "no")) |>
  kable()
```

Read the last two columns together. In bulk sediment aluminium accounts for roughly
half of why one sample holds more copper than another, which is what makes it a
grain-size control. Sieve the same sediment, and on comparable numbers of samples it
accounts for a few per cent at most. The rule asks for `r round(100 * smeta$al_r2_limit)`
per cent before aluminium counts as a control, so none of these is a near miss.
Dividing by it would not correct anything; it would add noise and call the result a
correction.

**The reason is that a sieved sample has already been grain-size corrected, by the
sieve.** Aluminium is a proxy for how muddy a sample is. Cut a sample down to
particles below 63 or 20 um and you have removed the mud-versus-sand difference
yourself, in the laboratory, before the chemistry starts. There is little texture
variation left for aluminium to track, so the correlation collapses. The normaliser
has not failed; it has been made redundant by the sample preparation, and what is
left of it is noise.

A second, rank-based check was run precisely so a low correlation could not be
blamed on the narrow range of the reference samples. It agrees on every group here.
Full workings: [Regression Normalisation](`r REFINED`/background-regression.html).

Sieved samples are not discarded. Their measurements, their background estimates
and their contamination classes are all published, the last of these because the
geo-accumulation index compares concentrations directly and needs no aluminium at
all. What a sieved sample does not get is a pristine verdict. The maps on the
element pages show bulk sediment for the related reason: it holds by far the most
data, and it is the only fraction the verdict speaks about.

## The verdict, and its cost

Two rules are published. The permissive one is a single test; the strict one
requires three independent criteria to agree.

```{r}
verdicts |>
  filter(cat == "bulk", reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  arrange(symbol) |>
  transmute(Element = sym_display[as.character(symbol)],
            Measurements = format(n, big.mark = "\u00a0"),
            `Can be judged` = sprintf("%s%%", pct_classifiable),
            `Judged pristine` = sprintf("%s%%", pct_ef),
            `Pristine, strict rule` = sprintf("%s%%", pct_strict)) |>
  kable()
```

The middle column is the finding that matters most, and it is a data gap rather
than a result. **Most samples cannot be judged at all**, because aluminium was
never measured alongside the metal.

```{r}
#| fig-height: 2.8
#| fig-cap: "The share of samples that can be judged falls away as you approach a fish farm."
cov <- coverage |>
  filter(axis == "distance to fish farm") |>
  mutate(band = factor(band, levels = band))
ggplot(cov, aes(band, pct_classifiable)) +
  geom_col(fill = "#CC6677") +
  geom_text(aes(label = paste0(pct_classifiable, "%")), vjust = -0.4, size = 3.3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = "Distance to the nearest fish farm", y = "Can be judged (%)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())
```

Aluminium coverage runs opposite to what the question needs: it is common in
offshore survey work and rare in near-cage monitoring. Almost none of the sediment
under a Norwegian fish farm can be given an aluminium-based verdict, which is why
the [aquaculture question](methods-pressure.qmd) is answered with a different tool.

Full workings: [Enrichment Factor](`r REFINED`/background-ef.html),
[Pristine Classification](`r REFINED`/background-pristine.html),
[Regression Normalisation](`r REFINED`/background-regression.html).
