---
title: "Aquaculture Pressure"
subtitle: "Whether fish farming shows up in the seabed, and how we made sure it was not an artefact"
---

{{< include _setup.qmd >}}

Norwegian salmon farming puts copper and zinc into the water: copper from net
antifouling, zinc from feed. If that reaches the seabed, sediment near a farm should
carry more of it than sediment far away.

Testing that turned out to be harder than it sounds, and the first answer we
published was wrong.

## The axis

Distance is measured to the nearest **marine fish farm**, not to any aquaculture
licence. The register also contains shellfish, land-based and freshwater sites,
which have no metal pressure and would blur the near band with sites that emit
nothing.

```{r}
#| output: asis
cat("```{mermaid}\n",
    "flowchart LR\n",
    '  A["Aquaculture register<br/>all licences"] --> F{"Marine fish<br/>farm?"}\n',
    '  F -->|no| X["Excluded:<br/>shellfish, land-based,<br/>freshwater"]\n',
    '  F -->|yes| D["Distance from each<br/>sediment site to the<br/>nearest one"]\n',
    '  D --> B1["under 1 km"]\n',
    '  D --> B2["1 to 5 km"]\n',
    '  D --> B3["5 to 20 km"]\n',
    '  D --> B4["over 20 km"]\n',
    "  style X fill:#eeeeee,stroke:#888\n",
    "  style D fill:#e8f0fe,stroke:#4477AA\n",
    "```\n", sep = "")
```

## The two things that fooled us

The far band is supposed to be the comparison: ordinary seabed, no farm nearby. It
is not, for two independent reasons.

::: {.callout-important}
## The comparison band was not clean, and not the same material

**It was full of monitored pollution.** More than half the copper measurements
beyond 20 km come from Norwegian programmes that exist to monitor contaminated
seabed, industry and sewage outfalls. Nobody surveys empty seabed at that
intensity. The far band was a harbour, not a reference.

**It was a different texture.** Sediment under a fish farm is roughly half as muddy
as offshore sediment, and mud carries metal. Compared against a muddy reference, a
sandy near-cage sample reads clean on grain size alone, before any farming is
considered.
:::

Both push the same way, and both flatten a real gradient. With them left in, the
data appeared to show that distance to a fish farm was a poor axis. That was a
property of the bands, not of the seabed.

## Controlling for both

```{r}
#| output: asis
cat("```{mermaid}\n",
    "flowchart TD\n",
    '  R["All measurements<br/>by distance band"] --> C1["Remove sites monitored<br/>for known pollution"]\n',
    '  C1 --> C2["Keep only samples in a<br/>common mud-content window"]\n',
    '  C2 --> G["Compare like with like"]\n',
    "  style R fill:#eeeeee,stroke:#888\n",
    "  style G fill:#d7ecd9,stroke:#1c5626\n",
    "```\n", sep = "")
```

Once both controls are applied, copper shows a clean gradient: highest under the
cages, falling with distance, lowest beyond 20 km.

```{r}
#| fig-height: 3.2
#| fig-cap: "Geo-accumulation index for copper in bulk sediment, after removing monitored pollution and matching mud content. Above zero means enriched relative to the regional background."
pm <- pressure |>
  filter(measure == "igeo_matched", cat == "bulk", symbol == "CU") |>
  mutate(band = factor(band, levels = c("<1km", "1-5km", "5-20km", ">20km")))
ggplot(pm, aes(band, value)) +
  geom_hline(yintercept = 0, colour = "#999") +
  geom_col(fill = "#4477AA", width = .6) +
  geom_text(aes(label = sprintf("%+.2f", value),
                vjust = ifelse(value >= 0, -0.4, 1.3)), size = 3.3) +
  scale_y_continuous(expand = expansion(mult = c(0.18, 0.18))) +
  labs(x = "Distance to the nearest fish farm", y = "Median Igeo (copper, bulk)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())
```

## Four controls, one answer

A single control can always be an accident of which samples it happened to keep.
Copper was tested four ways, on four different subsets of the data, and none of
them uses aluminium.

```{r}
controls |>
  filter(symbol == "CU", cat == "bulk", reliable) |>
  transmute(Control = control,
            `What it holds constant` = holds,
            `Samples near / far` = sprintf("%s / %s", format(n_near, big.mark = "\u00a0"),
                                           format(n_far, big.mark = "\u00a0")),
            `Copper near vs far` = detail) |>
  kable()
```

Four designs, four subsets of the data, the same direction. That is the argument.

::: {.callout-note}
## Zinc behaves differently, and we report that

```{r}
#| include: false
zn <- controls |> filter(symbol == "ZN", cat == "bulk")
zn_tex <- zn$detail[zn$control == "texture-matched"]
zn_mun <- zn$detail[zn$control == "same municipality"]
```

Zinc moves with distance under the texture control (`r zn_tex`) but loses it once
region is held constant (`r zn_mun`). Two controls disagreeing is a reason to read
zinc as sediment character rather than farm input, so we say so rather than
reporting the more favourable number.
:::

## What this does not show

- **Not a dose-response.** Larger farms do not straightforwardly mean more metal:
  large farms are sited in deeper, more exposed water, which disperses whatever
  they release. Depth and farm size are confounded.
- **Not a health assessment.** These are sediment concentrations, not exposure.
- **Not a verdict change.** The pristine classification never sees the aquaculture
  axis. That is deliberate: it is what allows this page to be an independent check
  on it rather than a restatement of it.

Full workings: [Pressure-Based Background](`r REFINED`/background-pressure.html),
[Pressure Controls](`r REFINED`/pressure-controls.html),
[Geo-Accumulation Index](`r REFINED`/background-igeo.html).
