---
title: "Estimation of Pristine Sediment"
subtitle: "Deciding whether one sample is at its background, and admitting when we cannot"
---

{{< include _setup.qmd >}}

A background is a number for a region. A **pristine verdict** is a judgement about
a single sample: is this one at its natural level, or above it?

The obstacle is grain size. Fine, muddy sediment naturally holds several times more
metal than coarse sand from the same seabed, with no pollution involved. Comparing
a raw concentration against a regional background therefore measures texture as
much as contamination.

## Two ways round it

```{r}
#| output: asis
cat("```{mermaid}\n",
    "flowchart TD\n",
    '  S["One sample"] --> Q{"Is aluminium<br/>measured?"}\n',
    '  Q -->|yes| EF["<b>Enrichment factor</b><br/>metal per aluminium,<br/>against an offshore reference"]\n',
    '  Q -->|no| I["<b>Geo-accumulation index</b><br/>metal against a regional<br/>background concentration"]\n',
    '  EF --> V["EF below 1<br/>= pristine"]\n',
    '  I --> W["Igeo at or below 0<br/>= unpolluted"]\n',
    "  style EF fill:#e8f0fe,stroke:#4477AA\n",
    "  style I fill:#fdf0d0,stroke:#7a5300\n",
    "```\n", sep = "")
```

**Aluminium is the grain-size correction.** It is a structural component of clay,
so it rises and falls with the mud content of a sample. Dividing the metal by it
removes the texture effect and leaves what is left over.

::: {.callout-warning}
## Iron is deliberately not used

Iron is the other conventional normaliser and would give far better coverage.
Norwegian fish feed contains iron, so iron near a farm is partly the very pressure
being measured. Using it would quietly subtract the signal. Iron is reported for
comparison and never used to issue a verdict.
:::

## Aluminium only works for three of the seven

A normaliser is only a grain-size control if it actually predicts the element.
Measured against the offshore reference, aluminium predicts cobalt, copper and zinc
in bulk sediment, and nothing else.

```{r}
elements |>
  filter(cat == "bulk") |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  arrange(symbol) |>
  transmute(Element = paste0(name, " (", sym_display[as.character(symbol)], ")"),
            `Aluminium predicts it` = ifelse(normalisable, "yes", "no"),
            `Verdict issued` = ifelse(has_verdict, "yes", "no"),
            Reason = vapply(as.character(symbol), capability_note, character(1))) |>
  kable()
```

Where aluminium fails, we issue no enrichment factor. The alternative would be to
divide by something that does not track texture and call the result a grain-size
control, which would be worse than saying nothing.

## And in bulk sediment only

The pristine verdict exists for bulk sediment alone, and aluminium is the reason.
Sieving is itself a grain-size correction, so aluminium is redundant in a sieved
sample and duly fails the test above. **This is not a shortage of sieved data.**

```{r}
pct_expl <- function(r2) ifelse(is.na(r2), NA_character_,
                        ifelse(100 * r2 < 0.1, "under 0.1%",
                               sprintf("%.1f%%", 100 * r2)))
elements |>
  filter(symbol %in% c("CU", "ZN", "CO"), al_tested) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         Fraction = factor(cat_label[cat], levels = cat_label[cat_order])) |>
  arrange(symbol, Fraction) |>
  transmute(Element = sym_display[as.character(symbol)],
            Fraction,
            `Measurements` = format(n, big.mark = "\u00a0"),
            `Samples with aluminium` = format(al_n, big.mark = "\u00a0"),
            `Variation aluminium explains` = pct_expl(al_r2)) |>
  kable()
```

Sieved copper has thousands of aluminium-paired samples and aluminium explains
almost nothing in them, against `r pct_expl(elements$al_r2[elements$symbol == "CU" & elements$cat == "bulk"])`
in bulk. The rule asks for `r round(100 * smeta$al_r2_limit)` per cent.

### Then why not judge the sieved samples without aluminium?

This is the right question, and the honest answer is that **nothing in the data
prevents it**. What prevents it is the definition. Our pristine verdict *is* an
enrichment factor: metal divided by aluminium, against an offshore reference. A
sieved sample does not need that correction, but it does not have the number
either, so it falls outside the rule rather than failing it.

The judgement that does work on a sieved sample, comparing its concentration
against a background for its own fraction, is already computed and published here
under a different name, the contamination class.

```{r}
verdicts |>
  filter(symbol %in% c("CU", "ZN", "CO"), !is.na(pct_igeo_class0)) |>
  mutate(symbol = factor(symbol, levels = elem_levels),
         Fraction = factor(cat_label[cat], levels = cat_label[cat_order])) |>
  arrange(symbol, Fraction) |>
  transmute(Element = sym_display[as.character(symbol)],
            Fraction,
            Measurements = format(n_igeo, big.mark = "\u00a0"),
            `Unpolluted (Igeo at or below 0)` = sprintf("%.1f%%", pct_igeo_class0),
            `Pristine verdict` = ifelse(is.na(pct_ef), "not issued",
                                        sprintf("%.1f%%", pct_ef))) |>
  kable()
```

```{r}
#| include: false
sv <- backgrounds |> filter(grepl("sieved", cat), reliable,
                            symbol %in% c("CU", "ZN", "CO"))
n_grp <- length(unique(paste(sv$symbol, sv$cat)))
n_p90 <- sum(sv$method == "offshore_p90")
n_mix <- sum(sv$method == "mixture")
```

The strict pristine rule is the enrichment factor plus two concentration criteria,
and those two survive without aluminium: an offshore 90th percentile exists for
`r if (n_p90 == n_grp) paste("all", n_grp) else paste(n_p90, "of the", n_grp)`
sieved groups here, and a distribution-mixture bound for `r n_mix` of them.

::: {.callout-note}
## An open decision, not a closed one

A concentration-based pristine verdict for the sieved fractions could be built from
what is already computed. It has not been, because it would be a different test
wearing the same word: the contamination class is measured against the offshore
median, and offshore is not the same as clean. Two verdicts that are not comparable
are worse than one verdict and one clearly labelled contamination class.
:::

## The verdict, and its cost

Two rules are published. The permissive one is a single test; the strict one
requires three independent criteria to agree.

```{r}
verdicts |>
  filter(cat == "bulk", reliable) |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  arrange(symbol) |>
  transmute(Element = sym_display[as.character(symbol)],
            Measurements = format(n, big.mark = "\u00a0"),
            `Can be judged` = sprintf("%s%%", pct_classifiable),
            `Judged pristine` = sprintf("%s%%", pct_ef),
            `Pristine, strict rule` = sprintf("%s%%", pct_strict)) |>
  kable()
```

The middle column is the finding that matters most, and it is a data gap rather
than a result. **Most samples cannot be judged at all**, because aluminium was
never measured alongside the metal.

```{r}
#| fig-height: 2.8
#| fig-cap: "The share of samples that can be judged falls away as you approach a fish farm."
cov <- coverage |>
  filter(axis == "distance to fish farm") |>
  mutate(band = factor(band, levels = band))
ggplot(cov, aes(band, pct_classifiable)) +
  geom_col(fill = "#CC6677") +
  geom_text(aes(label = paste0(pct_classifiable, "%")), vjust = -0.4, size = 3.3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(x = "Distance to the nearest fish farm", y = "Can be judged (%)") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())
```

Aluminium coverage runs opposite to what the question needs: it is common in
offshore survey work and rare in near-cage monitoring. Almost none of the sediment
under a Norwegian fish farm can be given an aluminium-based verdict, which is why
the [aquaculture question](methods-pressure.qmd) is answered with a different tool.

Full workings: [Enrichment Factor](`r REFINED`/background-ef.html),
[Pristine Classification](`r REFINED`/background-pristine.html),
[Regression Normalisation](`r REFINED`/background-regression.html).
