---
title: "All elements at a glance"
subtitle: "What is published for each of the seven, and why"
---

{{< include _setup.qmd >}}

Seven elements were targeted. Three can be given a full pristine verdict, four
cannot, and the reason is different in each case. Nothing here is a judgement about
how interesting an element is: the limits come from the measurements that exist.

```{r}
#| output: asis
n_full <- sum(sapply(unique(elements$symbol), capability) == "full")
n_part <- sum(sapply(unique(elements$symbol), capability) == "partial")
n_none <- sum(sapply(unique(elements$symbol), capability) == "none")
keyfig(
  list(n_full, "with a pristine verdict"),
  list(n_part, "with background only"),
  list(n_none, "not reported"),
  list(format(sum(elements$n), big.mark = "\u00a0"), "measurements behind it")
)
```

## The coverage matrix

```{r}
#| output: asis
tab <- elements |>
  group_by(symbol, name) |>
  summarise(n = first(n_elem), sites = first(n_sites_elem),
            cap = capability(first(symbol)),
            yr = paste0(first(year_min_elem), " to ", first(year_max_elem)),
            .groups = "drop") |>
  mutate(symbol = factor(symbol, levels = elem_levels)) |>
  arrange(symbol) |>
  transmute(Element = sprintf("[%s (%s)](element-%s.qmd)", name,
                              sym_display[as.character(symbol)],
                              tolower(as.character(symbol))),
            Measurements = format(n, big.mark = "\u00a0"),
            Sites = format(sites, big.mark = "\u00a0"),
            Years = yr,
            Published = chip(cap),
            Why = vapply(as.character(symbol), capability_note, character(1)))
kable(tab, escape = FALSE)
```

## Why some elements fall short

Two questions decide it, and the second one is asked only of bulk sediment: a
sieved sample was grain-size controlled by the sieve before it was analysed, so it
never needs aluminium in the first place.

```{r}
#| output: asis
cat("```{mermaid}\n",
    "flowchart TD\n",
    '  S["Seven target<br/>elements"] --> Q1{"Enough measurements<br/>above the detection limit?"}\n',
    '  Q1 -->|"no"| W1["<b>Se, Mo</b><br/>over half deleted<br/>below the LOQ"]\n',
    '  Q1 -->|"too few samples"| W2["<b>I</b><br/>15 measurements<br/>in total"]\n',
    '  Q1 -->|"yes"| Q2{"How is grain size<br/>controlled?"}\n',
    '  Q2 -->|"sieved: by the sieve"| V2["<b>Cu, Zn, Co, Mn</b><br/>background and<br/>pristine verdict"]\n',
    '  Q2 -->|"bulk: by aluminium"| Q3{"Does aluminium<br/>predict it?"}\n',
    '  Q3 -->|"no"| W3["<b>Mn</b><br/>background only,<br/>no enrichment factor"]\n',
    '  Q3 -->|"yes"| V["<b>Cu, Zn, Co</b><br/>background and<br/>pristine verdict"]\n',
    "  style V fill:#d7ecd9,stroke:#1c5626\n",
    "  style V2 fill:#d7ecd9,stroke:#1c5626\n",
    "  style W1 fill:#eeeeee,stroke:#888\n",
    "  style W2 fill:#eeeeee,stroke:#888\n",
    "  style W3 fill:#fdf0d0,stroke:#7a5300\n",
    "```\n", sep = "")
```

Manganese appears twice on purpose. In bulk it has no verdict because aluminium
was the grain-size control on offer and it does not work for manganese; in a
sieved sample the sieve supplies that control, so the question never arises.

Both rules were fixed before any result was inspected, and both are recorded as
frozen tables in the pipeline rather than recomputed at run time, so they cannot
move quietly under a rebuild.

## Background levels side by side

Bulk sediment, `r UNIT`. This is the offshore median, the most widely applicable of
the five estimates. Each element page carries all five.

```{r}
#| fig-height: 3.4
#| fig-cap: "Offshore median concentration, bulk sediment. Log scale: the seven elements span four orders of magnitude."
backgrounds |>
  filter(cat == "bulk", method == "offshore", reliable) |>
  left_join(elements |> filter(cat == "bulk") |> select(symbol, name), by = "symbol") |>
  mutate(lab = paste0(name, "\n(", sym_display[as.character(symbol)], ")"),
         lab = reorder(lab, value)) |>
  ggplot(aes(value, lab)) +
  geom_col(fill = "#4477AA") +
  geom_text(aes(label = fmt(value)), hjust = -0.15, size = 3.1) +
  scale_x_log10(expand = expansion(mult = c(0, 0.25))) +
  labs(x = paste0("Offshore median (", UNIT, ", log scale)"), y = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.y = element_blank())
```

## How much is pristine

Among the samples that can be judged at all.

```{r}
#| fig-height: 2.8
#| fig-cap: "Share of judgeable bulk samples classified pristine, by rule."
verdicts |>
  filter(cat == "bulk", reliable) |>
  select(symbol, `Permissive rule` = pct_ef, `Strict rule` = pct_strict) |>
  pivot_longer(-symbol, names_to = "rule", values_to = "pct") |>
  mutate(symbol = factor(sym_display[as.character(symbol)],
                         levels = sym_display[elem_levels])) |>
  ggplot(aes(symbol, pct, fill = rule)) +
  geom_col(position = position_dodge(width = .75), width = .65) +
  geom_text(aes(label = paste0(pct, "%")),
            position = position_dodge(width = .75), vjust = -0.4, size = 3) +
  scale_fill_manual(values = c(`Permissive rule` = "#4477AA", `Strict rule` = "#228833")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(x = NULL, y = "Pristine (%)", fill = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank(), legend.position = "top")
```

The two rules agree almost exactly, which is worth noting: the strict rule adds two
further criteria and barely changes the answer, so the verdict does not hinge on
where a single threshold was drawn.
