---
title: "All five sources at a glance"
subtitle: "What each archive brought, and how much of it survived"
---

{{< include _setup.qmd >}}

Nothing here was collected for this study. Five public archives were read as they
are published, and the measurements this site reports are what came through them.
They are not interchangeable: they cover different seas, different decades and
different elements, and they lose data at different stages and for different
reasons.

```{r}
#| output: asis
keyfig(
  list(nrow(sources), "archives"),
  list(format(sum(sources$n), big.mark = " "), "measurements reported"),
  list(format(smeta$n_sites, big.mark = " "), "sites"),
  list(paste0(min(sources$year_min), " to ", max(sources$year_max)), "sampling years")
)
```

## What each one contributes

```{r}
sources |>
  mutate(key = names(src_levels)[match(source, src_levels)],
         source = factor(source, levels = src_levels)) |>
  arrange(source) |>
  transmute(Source = sprintf("[%s](%s)", source, src_page(key)),
            Measurements = format(n, big.mark = " "),
            Share = sprintf("%.1f%%", pct),
            Sites = format(n_sites, big.mark = " "),
            Datasets = n_datasets,
            Elements = n_elements,
            Years = paste0(year_min, " to ", year_max)) |>
  kable(escape = FALSE)
```

```{r}
#| output: asis
top2 <- sources |> arrange(desc(pct)) |> head(2)
cat(sprintf("%s and %s account for %.0f measurements in every hundred between them. That is worth holding on to when reading any pooled result, and it is why the element pages report censoring per source rather than only in total.\n",
            top2$source[1], top2$source[2], sum(top2$pct)))
```

## Which elements each one carries

Counts are the measurements this site reports. A dash means the archive carries
no measurement of that element at all, which is a property of the archive and not
something the pipeline did. Each source page carries the archive's own totals
beside these.

```{r}
elem_cols <- elem_levels
source_elements |>
  mutate(cell = ifelse(in_archive, format(n_analysed, big.mark = " "), "-"),
         symbol = factor(symbol, levels = elem_cols),
         source = factor(source, levels = src_levels)) |>
  arrange(source, symbol) |>
  transmute(Source = sprintf("[%s](%s)", source, src_page(key)),
            Element = sym_display[as.character(symbol)], cell) |>
  pivot_wider(names_from = Element, values_from = cell) |>
  kable(escape = FALSE, align = c("l", rep("r", length(elem_cols))))
```

```{r}
#| include: false
carried <- source_elements |>
  group_by(symbol) |>
  summarise(n_src = sum(in_archive),
            who = paste(source[in_archive], collapse = " and "),
            .groups = "drop")
lone <- carried |> filter(n_src == 1)
all_five <- carried |> filter(n_src == nrow(sources))
```

```{r}
#| output: asis
if (nrow(lone)) {
  cat(sprintf(
    "The gaps are not evenly spread. %s %s carried by one archive alone (%s), which is most of the reason %s never reaches a reportable number. %s %s carried by all five.\n",
    paste(elem_name[as.character(lone$symbol)], collapse = " and "),
    if (nrow(lone) == 1) "is" else "are",
    paste(unique(lone$who), collapse = "; "),
    if (nrow(lone) == 1) "it" else "they",
    sub("^(.)", "\\U\\1", tolower(paste(elem_name[as.character(all_five$symbol)],
                                         collapse = " and ")), perl = TRUE),
    if (nrow(all_five) == 1) "is" else "are"))
}
```

## Which fraction each one measured

The fine fraction of a sediment carries most of the metal, so an archive that
sieves and one that does not are not measuring the same thing even when they
report the same element and the same seabed. Each source page carries its own
counts; the shares here are of everything that archive brought to the final
database.

```{r}
source_fractions |>
  mutate(cell = ifelse(n > 0, sprintf("%.1f%%", pct_of_source), "-"),
         cat = factor(frac_label[cat], levels = unname(frac_label)),
         source = factor(source, levels = src_levels)) |>
  arrange(source, cat) |>
  transmute(Source = sprintf("[%s](%s)", source, src_page(key)),
            Fraction = cat, cell) |>
  pivot_wider(names_from = Fraction, values_from = cell) |>
  kable(escape = FALSE, align = c("l", rep("r", length(frac_label))))
```

```{r}
#| include: false
one_frac <- source_fractions |>
  filter(pct_of_source >= 99) |>
  mutate(source = factor(source, levels = src_levels)) |>
  arrange(source)
```

```{r}
#| output: asis
if (nrow(one_frac))
  cat(sprintf("%s work in one fraction only. The others mix two or three, and every table on this site that touches grain size keeps the fractions apart rather than pooling them, because a bulk sample and a sieved one taken from the same seabed are not the same measurement.\n",
              paste(one_frac$source, collapse = " and ")))
```

## Aluminium, iron and organic carbon

None of these three is a measurand of this study. Aluminium is the grain-size
normaliser an enrichment factor is built on, iron carries much the same signal but
never enters a verdict here, and organic carbon binds metals and is read beside a
concentration. What matters is not how many of them an archive publishes but how
many of its reported measurements have one beside them, on the same sample and the
same fraction. That is what the shares below are.

```{r}
source_extras |>
  mutate(cell = ifelse(in_archive, sprintf("%.1f%%", pct_paired), "-"),
         name = factor(name, levels = unique(name)),
         source = factor(source, levels = src_levels)) |>
  arrange(source, name) |>
  transmute(Source = sprintf("[%s](%s)", source, src_page(key)),
            Parameter = name, cell) |>
  pivot_wider(names_from = Parameter, values_from = cell) |>
  kable(escape = FALSE, align = c("l", "r", "r", "r"))
```

```{r}
#| include: false
alu <- source_extras |> filter(symbol == "AL") |> arrange(desc(pct_paired))
```

The spread in the aluminium column is the single widest gap between these five
archives: `r alu$source[1]` pairs `r sprintf("%.1f%%", alu$pct_paired[1])` of what
it reports with an aluminium, `r alu$source[nrow(alu)]` only
`r sprintf("%.1f%%", alu$pct_paired[nrow(alu)])`. An enrichment factor cannot be
computed without one, so that column is a ceiling on how much of each archive can
ever carry an enrichment-based verdict. It is also what sits behind the
classifiable share on
[How a pristine sample is decided](methods-pristine.qmd): nearly all the sampling
close to a fish farm comes from the archive in this table with the least
aluminium beside it.

## How much survives each step

Every archive loses data, and the shape of the loss is the interesting part. The
same six steps are drawn for each, as a share of what the parse found.

```{r}
#| fig-height: 3.8
#| fig-cap: "Target-element measurements surviving each step, as a share of what the parse found."
source_flow |>
  mutate(stage = factor(stage_label[stage], levels = unname(stage_label)),
         source = factor(source, levels = src_levels)) |>
  ggplot(aes(stage, pct_of_pilot, colour = source, group = source)) +
  geom_line(linewidth = .7) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = src_palette) +
  scale_y_continuous(limits = c(0, 100),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Surviving (%)", colour = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank(), legend.position = "top",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r}
#| include: false
lost_merge <- source_flow |>
  filter(stage == "merged", pct_of_prev < 99) |>
  pull(source)
```

Only the cleaning step bites everywhere. The merge step is nearly free for three
of the five and costs `r if (length(lost_merge)) paste(lost_merge, collapse = " and ") else "none of them"`
a visible share, because that is where the same cruise reported into two archives
is reduced to one copy, and the copy that is kept is the one from the archive
closest to the measurement.

## Where each one sampled

One cell is 0.1 degrees, roughly 11 km, the same grid the element maps use.
Colour is the archive, and size is how many sites the cell holds. The fraction
control narrows it to the cells where that fraction was actually sampled, which is
how the sieved and unsieved parts of this study can be seen apart. Nothing here is
a concentration: this map answers where and how the sampling happened, not what
was found.

```{r}
#| include: false
smap <- read.csv(file.path(sdir, "summary_source_map.csv"),
                 stringsAsFactors = FALSE)
smap <- smap[, c("source", "key", "lat", "lon", "n_sites", "n", "n_elements",
                 "year_min", "year_max", "depth_p50",
                 "n_bulk", "n_sieved63", "n_sieved20")]
ojs_define(srcmap = smap)
ojs_define(src_order = data.frame(source = unname(src_levels),
                                  key = names(src_levels),
                                  colour = unname(src_palette[unname(src_levels)]),
                                  stringsAsFactors = FALSE))
```

```{=html}
<link rel="stylesheet" href="https://unpkg.com/leaflet@1.9.4/dist/leaflet.css" />
```

```{ojs}
//| echo: false
L = require("https://unpkg.com/leaflet@1.9.4/dist/leaflet.js")
allcells = transpose(srcmap)
srcs = transpose(src_order)
```

```{ojs}
//| echo: false
// All five on at once is the default, because the question the map answers is
// where they overlap and where only one archive has been.
viewof shown = Inputs.checkbox(new Map(srcs.map(o => [o.source, o.source])),
                               { value: srcs.map(o => o.source),
                                 label: "Archives" })
```

```{ojs}
//| echo: false
// A filter rather than a second colour: colour already means the archive here,
// and a cell shown under one fraction is the same cell, not a recount. The
// counts in the popup are the whole cell, as they are everywhere else.
viewof frac = Inputs.radio(new Map([["Any fraction", "any"], ["Bulk", "n_bulk"],
                                    ["Sieved <63 um", "n_sieved63"],
                                    ["Sieved <20 um", "n_sieved20"]]),
                           { value: "any", label: "Fraction" })
```

```{ojs}
//| echo: false
cells = allcells.filter(d => shown.includes(d.source) &&
                            (frac === "any" || d[frac] > 0))
colour_of = new Map(srcs.map(o => [o.source, o.colour]))
```

```{ojs}
//| echo: false
srcMapDiv = {
  const div = html`<div id="map"></div>`;
  yield div;
  const map = L.map(div, { zoomControl: true });
  L.tileLayer("https://tile.openstreetmap.org/{z}/{x}/{y}.png", {
    attribution: "&copy; OpenStreetMap contributors",
    maxZoom: 12
  }).addTo(map);

  // The home view is the extent of every archive, not of whatever is ticked, so
  // turning one off does not move the map under the reader. The floor of 3 is
  // the same one the element maps use: Mercator stretches the far north, and
  // fitting 36 to 81 degrees honestly opens on a world map.
  const lats = allcells.map(d => d.lat), lons = allcells.map(d => d.lon);
  const home = L.latLngBounds([Math.min(...lats), Math.min(...lons)],
                              [Math.max(...lats), Math.max(...lons)]);
  const goHome = () => {
    map.invalidateSize();
    map.fitBounds(home, { padding: [24, 24], maxZoom: 6 });
    if (map.getZoom() < 3) map.setView(home.getCenter(), 3);
  };
  goHome();

  const HomeControl = L.Control.extend({
    options: { position: "topleft" },
    onAdd() {
      const btn = L.DomUtil.create("a", "leaflet-home-button");
      btn.href = "#";
      btn.title = "Reset the view";
      btn.setAttribute("aria-label", "Reset the view");
      btn.innerHTML = '<i class="bi bi-house-door" aria-hidden="true"></i>';
      L.DomEvent.on(btn, "click", e => { L.DomEvent.stop(e); goHome(); });
      return btn;
    }
  });
  new HomeControl().addTo(map);

  // What the cell holds in each fraction, in every popup, whichever filter is on.
  const fracline = d => [["n_bulk", "bulk"], ["n_sieved63", "sieved <63 um"],
                         ["n_sieved20", "sieved <20 um"]]
    .filter(([k]) => d[k] > 0).map(([k, l]) => `${l} ${d[k]}`).join(", ");

  // Drawn in reverse order of size so the sparse archives are not buried under
  // the dense ones: 4Demon has 25 cells against ICES-DOME's 2 679.
  const order = srcs.map(o => o.source)
    .filter(s => shown.includes(s))
    .sort((a, b) => cells.filter(d => d.source === b).length -
                    cells.filter(d => d.source === a).length);

  for (const s of order) {
    for (const d of cells.filter(d => d.source === s)) {
      L.circleMarker([d.lat, d.lon], {
        radius: Math.min(7, 2 + Math.sqrt(d.n_sites)),
        color: colour_of.get(s),
        weight: 0.5,
        fillColor: colour_of.get(s),
        fillOpacity: 0.65
      }).bindPopup(
        `<b>${s}</b><br/>${d.n_sites} site${d.n_sites === 1 ? "" : "s"}` +
        `<br/>${d.n} measurements, ${d.n_elements} element${d.n_elements === 1 ? "" : "s"}` +
        `<br/>${d.year_min} to ${d.year_max}` +
        (d.depth_p50 === null || isNaN(d.depth_p50) ? "" : `<br/>median depth ${d.depth_p50} m`) +
        `<br/>${fracline(d)}`
      ).addTo(map);
    }
  }

  const legend = L.control({ position: "bottomright" });
  legend.onAdd = () => {
    // the element maps' own legend classes, so the two read as one object
    const box = L.DomUtil.create("div", "legend");
    box.innerHTML = '<div class="legend-title">Archive</div>' + order.map(s =>
      `<div class="legend-row"><span class="legend-swatch" ` +
      `style="background:${colour_of.get(s)}"></span>${s}</div>`
    ).join("") +
    `<div class="legend-note">Cells are 0.1 degrees. Size is the number of sites. ` +
    `A fraction filter shows the cells that hold at least one measurement in it.</div>`;
    return box;
  };
  legend.addTo(map);
}
```

Where the cells sit on top of each other is where two archives sampled the same
water, and that is what the merge step has to resolve. The North Sea and the
Skagerrak are the crowded part; the Barents Sea is almost entirely one archive.
